---
title: "Combine_whale_zoop"
author: "Sarah Roberts"
date: "2023-03-08"
output: html_document
---

#libraries
```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(sp)
library(sf)
library(spatialEco)
library(rnaturalearth)
library(olsrr)
library(party)
library(gridExtra)
library(gjam)
library(lessR)
library(ggthemes)
library(wesanderson)
library(ggpubr)
library(pROC)
library(ggsci)
library(gridExtra)
library(cowplot)
library(patchwork)
library(lessR)
library(ggcorrplot)
world <- ne_countries(scale = "medium", returnclass = "sf")

library(mapdata)
library(maps)

usa <- map_data('usa')

remove_outliers <- function(df, cols = names(df)) {
  for (col in cols) {
    df <- df[!outliers(df[[col]]),]
  }
  df
}

sumna <- function(x){
  #acts like sum(na.rm=T) but returns NA if all are NA
  if(!all(is.na(x))) return(sum(x, na.rm=T))
  if(all(is.na(x))) return(NA)
}

r2_general <-function(preds,actual){ 
  return(1- sum((preds - actual) ^ 2)/sum((actual - mean(actual))^2))
}

RMSE_func <- function(preds, actual){
  return(sqrt(mean((actual - preds)^2)))
}

theme_Publication <- function(base_size=12, base_family="Arial") {

      (theme_foundation(base_size=base_size, base_family=base_family)
       + theme(plot.title = element_text(face = "bold",
                                         size = rel(1.2), hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               legend.key.size= unit(0.5, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))
      
}

theme_Publication <- function(base_size=12, base_family="Arial") {

      (theme_foundation(base_size=base_size, base_family=base_family)
       + theme(plot.title = element_text(hjust = 0.5),
               text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(), 
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "bottom",
               legend.direction = "horizontal",
               legend.key.size= unit(0.5, "cm"),
               legend.margin = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0")
          ))
      
}

pal <- wes_palette("Zissou1", 50, type = "continuous")

'%!in%' <- function(x,y)!('%in%'(x,y))

```

This is reading in the text file I made from the survey shp file. You can also read in the shapefile and use those polygons (then calculate density per segment area)
```{r}
whale <- read.table("whale_data/baleen_segments.txt", header = TRUE)
whale <- whale %>% dplyr::filter(ModelID == 28 | ModelID == 29) #Humpback and Fin
```

28 - Humpback
29 - Fin 
33 - Sei 

Make the date column a date and separate out 

```{r}
whale$date <- as.Date(whale$StartTime, format = "%m/%d/%Y")
whale <- whale %>% mutate(day = day(date), month = month(date), year = year(date))
```

```{r}


ggplot(data = world) + geom_sf() +geom_point(data = subset(whale, whale$Individu_1 > 0), aes(x = X_mid, y = Y_mid, colour = as.factor(ModelID))) + facet_wrap(~year)+
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)

ggplot(data = world) + geom_sf() +geom_point(data = subset(whale, whale$Individu_1 > 0), aes(x = X_mid, y = Y_mid, colour = as.factor(year))) + facet_wrap(~month)+
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)

whale %>% filter(Individu_1 > 0) %>% ggplot(aes(x = month)) + geom_histogram(aes(group = as.factor(ModelID), fill = as.factor(ModelID)), position = "dodge") + facet_wrap(~year)
```

#ecomon data 
load in the data 
```{r}
ecomon <- read.csv("ecomon_data/EcoMon_Plankton_Data_v3_7-Data.csv")
```


First lets look at the structure of the data and convert date to a date column
```{r}
str(ecomon)

#convert data and create year, month, day columns: 

ecomon$date <- as.Date(ecomon$date, format =  "%d-%b-%y") #b here is the code for month when its words lol
ecomon <- ecomon %>% mutate(day = day(date), month = month(date), year = year(date))
ecomon$haulid <- paste(ecomon$cruise_name, "_", ecomon$station, "_", ecomon$depth)
```


#overlap grid 
first lets try out a simple grid of overlapping points (need to think about the different data structures, but for now lets forge ahead)
```{r}
# make an object the size and shape of the output you want
globe_bb <- matrix(c(-84,  20,
                      -84,  50,
                      -52, 50,
                     -52, 20,
                     -84,  20), byrow = TRUE, ncol = 2) %>%
  list() %>% 
  st_polygon() %>% 
  st_sfc(., crs = 4326)

#checkit
ggplot(data = world) + geom_sf() +geom_sf(data = globe_bb) + geom_point(data = whale, aes(x = X_mid, y = Y_mid)) + geom_point(data = ecomon, aes(x = lon, y = lat), colour = "red") + 
    coord_sf(xlim=c(-90, -20), ylim=c(20,60), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


# generate grid of 128 x 128 tiles (each .5 x .5 degrees)
globe_grid_18x18 <- st_make_grid(globe_bb, n = c(64, 64), 
                                 crs = 4326, what = 'polygons') %>%
  st_sf('geometry' = ., data.frame('ID' = 1:length(.)))

#checkit
ggplot(data = world) + geom_sf() +geom_sf(data = globe_grid_18x18) +
    coord_sf(xlim=c(-90, -50), ylim=c(20,60), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

zoop_table <- ecomon

#now sample grid ID for zoop table and whale table 
zoop_spat <- ecomon


coordinates(zoop_spat) <- ~ lon + lat
zoop_spat <- st_as_sf(zoop_spat)

zoop_spat <- st_set_crs(zoop_spat, 4326)
grid <- globe_grid_18x18


zoop_extract <- point.in.poly(zoop_spat, grid)
zoop_extract_2 <- as.data.frame(zoop_extract)


whale_spat <- whale


coordinates(whale_spat) <- ~ X_mid + Y_mid
whale_spat <- st_as_sf(whale_spat)
whale_spat <- st_set_crs(whale_spat, 4326)
grid <- globe_grid_18x18


whale_extract <- point.in.poly(whale_spat, grid)
whale_extract_2 <- as.data.frame(whale_extract)



#change to grid id 
whale_extract_2$gridID <- whale_extract_2$ID
zoop_extract_2$gridID <- zoop_extract_2$ID



#group to get rid of duplicates (there are times when a point is overlapping two grid cells)
#group by 
zoop_extract_2_2 <- zoop_extract_2[!duplicated(zoop_extract_2[c("haulid", "coords.x1","coords.x2")]),] 

whale_extract_2_2 <- whale_extract_2[!duplicated(whale_extract_2[c("ID.x", "coords.x1","coords.x2")]),] 

zoop_extract_2_2 <- zoop_extract_2_2 %>% group_by(year, month, gridID) %>% dplyr::summarise_if(is.numeric, mean, na.rm = TRUE)

zoop_extract_2_2[zoop_extract_2_2 == "NaN"] = NA

whale_extract_2_2 <- whale_extract_2_2 %>% group_by(ModelID, year, month, ID.y) %>% dplyr::summarise(Individu_1 = sum(Individu_1), length = sum(SegmentLen), coords.x1 = mean(coords.x1), coords.x2 = mean(coords.x2), month = mean(month))
  
#need to take the sum of Individu_1 and sum of survey lengths to get a new abd per km 
whale_extract_2_2$AbdPerArea <- whale_extract_2_2$Individu_1/whale_extract_2_2$length
  


both <- left_join(whale_extract_2_2, zoop_extract_2_2, by = c("year" = "year", "month" = "month", "ID.y" = "gridID"))



both <- both %>% filter(ammspp_10m2 >-1)

#add distance to shore (to separate ecotypes later on)
both_spat <- both
coordinates(both_spat) <- ~ coords.x1.x + coords.x2.x
proj4string(both_spat) <- CRS("+init=epsg:4326")
both_spat <- spTransform(both_spat, "+proj=utm +zone=19 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
library(rgeos)

countries <- as_Spatial(world)
proj4string(countries) <- CRS("+init=epsg:4326")




countries_sub <- subset(countries, countries@data[["subregion"]] == "Northern America")
countries_sub <- spTransform(countries_sub, "+proj=utm +zone=19 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")

plot(both_spat, col = "red")
plot(countries_sub, add = T)

both$dist_to_shore <- apply(gDistance(both_spat, countries_sub,byid=TRUE),2,min)
both$dist_to_shore <- both$dist_to_shore/1000

write.csv(both, "whales_and_zoop_monthly.csv")

```

#correlations 
find out which fish are correlated with the whales 

```{r}
both <- both %>% pivot_wider(names_from = ModelID, values_from = AbdPerArea)
both <- both %>% dplyr::rename("Humpback" = "28", "Fin" = "29")


humpback <- both %>% dplyr::filter(!is.na(Humpback)) %>% select(-Fin)

fin <- both %>% dplyr::filter(!is.na(Fin))  %>% select(-Humpback)

both_subset1 <- humpback %>% select_if(is.numeric) %>% drop_na() 
ycor <- cor(both_subset1, method = "spearman")
ycor <-  as.data.frame(ycor) 
rownames(ycor) <- colnames(ycor)

ycor <- ycor %>% select(Humpback)
ycor$cor <- rownames(ycor)
ycor %>% arrange(-Humpback)
```


It looks like the top ones are pseudo_10m2, myoaen_10m2,calfin_10m2, acarlong_10m2, and mlucens_10m2

those are Pseudocalanus (copepod), Myoxocephalus aenaeus (scuplin), Calanus finmarchicus (copepod), Acartia longiremis (copepod) and Metridia lucens (copepod)

The literature says they feed on capelin (Mallotus villosus), herring (Clupea harengus), sand lance (Ammodytes spp.), and krill (Meganyctiphanes norvegica) https://onlinelibrary-wiley-com.libproxy.lib.unc.edu/doi/10.1111/gcb.16225
Problem is the zoop data is fish larvae, not the actual fish which is what humpbacks feed on in addition to plankton 



Lets take a look at the overlap of those copepods and humpbacks 

##hump eda
```{r}
specs <- c("pseudo_10m2", "myoaen_10m2","calfin_10m2", "acarlong_10m2", "mlucens_10m2", "cluhar_10m2", "ammspp_10m2", "megan_10m2", "euph_10m2")
specs1 <- c("pseu_copepod", "scuplin", "calanus", "arc_copepod", "mlcu_copepod", "herring", "sand_lance", "meg_krill", "euph_krill")

humpback <- humpback %>% mutate(pseu_copepod = pseudo_10m2, scuplin = myoaen_10m2, calanus = calfin_10m2, arc_copepod = acarlong_10m2, mlcu_copepod =  mlucens_10m2, herring = cluhar_10m2, sand_lance = ammspp_10m2, meg_krill = megan_10m2, euph_krill = euph_10m2)

dat <- humpback %>% pivot_longer(cols = specs1, names_to = "zoop", values_to = "zoop_m2")


dat <- dat[colnames(dat) %in% c("Humpback", "zoop", "zoop_m2", "coords.x1.x", "coords.x2.x", "month", "year")]

for (i in 1:length(specs1)){
  use <- specs1[i]
  dat1 <- dat %>% dplyr::filter(zoop == use)
  dat1 <- dat1 %>% dplyr::filter(zoop_m2 >0 & Humpback > 0 )
print(ggplot(data = world) + geom_sf() + geom_point(data = dat1, aes(x = coords.x1.x, y = coords.x2.x), size = .4) +
    coord_sf(xlim=c(-78, -62), ylim=c(35,45), expand = TRUE)  + facet_wrap(~month) + ggtitle(label = paste(use, nrow(dat1))))
}


for (i in 1:length(specs1)){
  use <- specs1[i]
  dat1 <- dat %>% dplyr::filter(zoop == use)
  dat1 <- dat1 %>% dplyr::filter(zoop_m2 >0 & Humpback > 0 )
print(ggplot(data = world) + geom_sf() + geom_point(data = dat1, aes(x = coords.x1.x, y = coords.x2.x), size = .4) +
    coord_sf(xlim=c(-78, -62), ylim=c(35,45), expand = TRUE)  + facet_wrap(~year) + ggtitle(label = paste(use, nrow(dat1))))
}
```
```{r}
for (i in 1:length(specs1)){
  use <- specs1[i]
  dat1 <- dat %>% dplyr::filter(zoop == use)
  dat1 <- dat1 %>% dplyr::filter(zoop_m2 >0 & Humpback > 0 )
print(ggplot(data = dat1) + geom_histogram(aes(x = month)) + facet_wrap(~year) + ggtitle(label = paste(use, nrow(dat1))))
}
```


##fin eda
Same thing with fin whales 

```{R}
both_subset1 <- fin %>% select_if(is.numeric) %>% drop_na() 
ycor <- cor(both_subset1, method = "spearman")
ycor <-  as.data.frame(ycor) 
rownames(ycor) <- colnames(ycor)

ycor <- ycor %>% select(Fin)
ycor$cor <- rownames(ycor)
ycor %>% arrange(-Fin)
```


It looks like the top ones are calfin_10m2, mlucens_10m2, pseudo_10m2, myoaen_10m2,calfin_10m2, acarlong_10m2, and mlucens_10m2, sebspp_100m3, euph_10m2

those are Pseudocalanus (copepod), Myoxocephalus aenaeus (scuplin), Calanus finmarchicus (copepod), Acartia longiremis (copepod) and Metridia lucens (copepod)

The literature says they feed on capelin (Mallotus villosus), herring (Clupea harengus), sand lance (Ammodytes spp.), and krill (Meganyctiphanes norvegica) https://onlinelibrary-wiley-com.libproxy.lib.unc.edu/doi/10.1111/gcb.16225
Problem is the zoop data is fish larvae, not the actual fish which is what humpbacks feed on in addition to plankton 



Lets take a look at the overlap of those copepods and humpbacks 


```{r}
specs <- c("pseudo_10m2", "myoaen_10m2","calfin_10m2", "acarlong_10m2", "mlucens_10m2", "cluhar_10m2", "ammspp_10m2", "megan_10m2", "euph_10m2")
specs1 <- c("pseu_copepod", "scuplin", "calanus", "arc_copepod", "mlcu_copepod", "herring", "sand_lance", "meg_krill", "euph_krill")

fin <- fin %>% mutate(pseu_copepod = pseudo_10m2, scuplin = myoaen_10m2, calanus = calfin_10m2, arc_copepod = acarlong_10m2, mlcu_copepod =  mlucens_10m2, herring = cluhar_10m2, sand_lance = ammspp_10m2, meg_krill = megan_10m2, euph_krill = euph_10m2)

dat <- fin %>% pivot_longer(cols = specs1, names_to = "zoop", values_to = "zoop_m2")


dat <- dat[colnames(dat) %in% c("Fin", "zoop", "zoop_m2", "coords.x1.x", "coords.x2.x", "month", "year")]

for (i in 1:length(specs1)){
  use <- specs1[i]
  dat1 <- dat %>% dplyr::filter(zoop == use)
  dat1 <- dat1 %>% dplyr::filter(zoop_m2 >0 & Fin > 0 )
print(ggplot(data = world) + geom_sf() + geom_point(data = dat1, aes(x = coords.x1.x, y = coords.x2.x), size = .4) +
    coord_sf(xlim=c(-78, -62), ylim=c(35,45), expand = TRUE)  + facet_wrap(~month) + ggtitle(label = paste(use, nrow(dat1))))
}


for (i in 1:length(specs1)){
  use <- specs1[i]
  dat1 <- dat %>% dplyr::filter(zoop == use)
  dat1 <- dat1 %>% dplyr::filter(zoop_m2 >0 & Fin > 0 )
print(ggplot(data = world) + geom_sf() + geom_point(data = dat1, aes(x = coords.x1.x, y = coords.x2.x), size = .4) +
    coord_sf(xlim=c(-78, -62), ylim=c(35,45), expand = TRUE)  + facet_wrap(~year) + ggtitle(label = paste(use, nrow(dat1))))
}
```
```{r}
for (i in 1:length(specs1)){
  use <- specs1[i]
  dat1 <- dat %>% dplyr::filter(zoop == use)
  dat1 <- dat1 %>% dplyr::filter(zoop_m2 >0 & Fin > 0 )
print(ggplot(data = dat1) + geom_histogram(aes(x = month)) + facet_wrap(~year) + ggtitle(label = paste(use, nrow(dat1))))
}
```

#Questions
1 Who are humpbacks and fin associated with? 
  Are their distributions explained by zoops? 
2 Has association changed over time? 
  Start with highly co-occuring species (maybe 20 or 50)
  Run gjam to get out environmental covariance or co-occurence or deviance explained 
  look at dendrograms over time (maybe start with just the summer)
3 Has phenology changed? 
  Are they switching between prey because the prey have moved out of the area earlier? 
4 is there diet overlap? Niche specificity? 















#Fish
do we have any sand lance, herring, capelin info in the trawls? 

```{r}
load("NEFSC_BTS_2021_all_seasons.RData")
load("stratum.area.RData")
fish_table <- survey$survdat
#this is monthly cpue (wtcpue)

# sum different sexes of same spp together
#okay so now we know that the biomass column was given to them as the calibrated cpue
fish_table <- fish_table %>% 
  group_by(YEAR, SEASON, EST_TOWDATE, LAT, LON, CRUISE6, DEPTH, STATION, STRATUM, SVSPP, COMNAME) %>% 
  dplyr::summarise(wtcpue = sum(BIOMASS), abd = sum(ABUNDANCE), BOTTEMP = mean(BOTTEMP), BOTSALIN = mean(BOTSALIN), SURFTEMP = mean(SURFTEMP), SURFSALIN = mean(SURFSALIN)) 

neus_strata <- stratum.area %>% 
  dplyr::select(STRATUM, STRATUM_AREA) %>% 
  mutate(STRATUM = as.integer(STRATUM)) %>%
  distinct()

fish_table <-  left_join(fish_table, neus_strata, by = c("STRATUM" = "STRATUM"))

fish_table <- filter(fish_table, !is.na(STRATUM_AREA))

fish_table <- fish_table %>%
  mutate(
    # Create a unique haulid
    haulid = paste(formatC(CRUISE6, width=6, flag=0), formatC(STATION, width=3, flag=0), formatC(STRATUM, width=4, flag=0), sep='-'),  
    # Calculate stratum area where needed (use convex hull approach)
    # convert square nautical miles to square kilometers
    STRATUM_AREA = STRATUM_AREA * 3.429904) %>% 
  dplyr::rename(year = YEAR,
         spp = COMNAME,
         date = EST_TOWDATE,
         lat = LAT, 
         lon = LON, 
         depth = DEPTH,
         stratum = STRATUM, 
         stratumarea = STRATUM_AREA) %>%
  filter(
    # remove unidentified spp and non-species
    spp != "" | !is.na(spp), 
    !grepl("EGG", spp), 
    !grepl("UNIDENTIFIED", spp)) %>%
  group_by(haulid, stratum, stratumarea, year, date, lat, lon, depth, spp, SEASON) %>% 
  dplyr::summarise(wtcpue = sumna(wtcpue), BT = mean(BOTTEMP), BSAL = mean(BOTSALIN), SST = mean(SURFTEMP), SSAL = mean(SURFSALIN)) %>% 
  # add temporary region column (this will be replaced with seasonal name)
  mutate(region = "Northeast US") %>% 
  dplyr::select(region, haulid, year, date, lat, lon, stratum, stratumarea, depth, spp, wtcpue, SEASON, BT, SST, BSAL, SSAL) %>% 
  ungroup()


fish_table <- fish_table %>% 
  drop_na(lat,lon)

fish_table$month <- month(as.POSIXlt(fish_table$date))


fish_table_wide <- pivot_wider(fish_table, names_from = spp, values_from = wtcpue)
 #with more than 20% NA

fish_table <- fish_table_wide

#convert NA to 0 
fish_table[, 15:ncol(fish_table)][is.na(fish_table[, 15:ncol(fish_table)])] <- 0

colnames(fish_table) <- gsub(pattern = ' ', replacement = ".", x = colnames(fish_table))
fish_table <- fish_table %>% dplyr::select(-SEASON)


fish_test <- fish_table %>% 
  drop_na(lat,lon)

fish_table <- fish_test

#now sample grid ID for fish table and dolphin table 
fish_spat <- fish_test


coordinates(fish_spat) <- ~ lon + lat
fish_spat <- st_as_sf(fish_spat)

fish_spat <- st_set_crs(fish_spat, 4326)
grid <- globe_grid_18x18


fish_extract <- point.in.poly(fish_spat, grid)
fish_extract_2 <- as.data.frame(fish_extract)



fish_extract_2$gridID <- fish_extract_2$ID



#group to get rid of duplicates (there are times when a point is overlapping two grid cells)
#group by 
fish_extract_2_2 <- fish_extract_2[!duplicated(fish_extract_2[c("haulid", "coords.x1","coords.x2")]),] 

fish_extract_2_2 <- fish_extract_2_2 %>% group_by(year, month, gridID) %>% dplyr::summarise_if(is.numeric, mean, na.rm = TRUE)

fish_extract_2_2[fish_extract_2_2 == "NaN"] = NA

both_fish <- left_join(whale_extract_2_2, fish_extract_2_2, by = c("year" = "year", "month" = "month", "ID.y" = "gridID"))


both_fish <- both_fish %>% filter(ATLANTIC.CROAKER >-1)
both_fish <- both_fish %>% ungroup() %>% drop_na(SST,SSAL, depth)

#add distance to shore (to separate ecotypes later on)
both_spat <- both_fish
coordinates(both_spat) <- ~ coords.x1.x + coords.x2.x
proj4string(both_spat) <- CRS("+init=epsg:4326")
both_spat <- spTransform(both_spat, "+proj=utm +zone=19 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
library(rgeos)

countries <- as_Spatial(world)
proj4string(countries) <- CRS("+init=epsg:4326")




countries_sub <- subset(countries, countries@data[["subregion"]] == "Northern America")
countries_sub <- spTransform(countries_sub, "+proj=utm +zone=19 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")

plot(both_spat, col = "red")
plot(countries_sub, add = T)

both_fish$dist_to_shore <- apply(gDistance(both_spat, countries_sub,byid=TRUE),2,min)
both_fish$dist_to_shore <- both_fish$dist_to_shore/1000

write.csv(both_fish, "both_fish.csv")

```

#correlations 
find out which fish are correlated with the whales 

```{r}
both_subset1 <- both_fish %>%select_if(is.numeric) %>% drop_na() 
ycor <- cor(both_subset1, method = "spearman")
ycor <-  as.data.frame(ycor) 
rownames(ycor) <- colnames(ycor)

ycor <- ycor %>% select(AbdPerArea)
ycor$cor <- rownames(ycor)
ycor %>% arrange(-AbdPerArea)
```
haddock, herring, cod, mackerel, sculpin, plaice 

```{r}

specs1 <- c("HADDOCK", "ATLANTIC.HERRING", "ATLANTIC.COD", "ATLANTIC.MACKEREL", "LONGHORN.SCULPIN", "AMERICAN.PLAICE", "NORTHERN.SAND.LANCE")

dat <- both_fish %>% pivot_longer(cols = specs1, names_to = "fish", values_to = "fish_abd")


dat <- dat[colnames(dat) %in% c("AbdPerArea", "fish", "fish_abd", "coords.x1.x", "coords.x2.x", "month")]

for (i in 1:length(specs1)){
  use <- specs1[i]
  dat1 <- dat %>% dplyr::filter(fish == use)
  dat1 <- dat1 %>% dplyr::filter(fish_abd >0 & AbdPerArea > 0 )
print(ggplot(data = world) + geom_sf() + geom_point(data = dat1, aes(x = coords.x1.x, y = coords.x2.x), size = .4) +
    coord_sf(xlim=c(-78, -62), ylim=c(35,45), expand = TRUE)  + facet_wrap(~month) + ggtitle(label = paste(use, nrow(dat1))))
}
```

it looks like they don't overlap with american sand lance or northern sand lance but that is because those two species are not erally captured in the trawls. 
I think the only usable one here is herring - the problem is the timing of the surveys - don't include summer and summer is when they are feeding in the gulf of main. Lets stick to zoop data for now 
