---
title: "Figures"
author: "Sarah Roberts"
date: "2024-01-09"
output: html_document
---

#libraries
```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggpmisc)

library(sp)
library(sf)
library(spatialEco)
library(rnaturalearth)
library(olsrr)
library(party)
library(gridExtra)
library(gjam)
library(lessR)
library(ggthemes)
library(wesanderson)
library(ggpubr)
library(pROC)
library(ggsci)
library(gridExtra)
library(cowplot)
library(patchwork)
library(lessR)
library(ggcorrplot)
library(tidytext)
library(geosphere)
library(mgcv)
library(fitdistrplus)
library(units)
library(ggh4x)

world <- ne_countries(scale = "medium", returnclass = "sf")

library(mapdata)
library(maps)

usa <- map_data('usa')

remove_outliers <- function(df, cols = names(df)) {
  for (col in cols) {
    df <- df[!outliers(df[[col]]),]
  }
  df
}

sumna <- function(x){
  #acts like sum(na.rm=T) but returns NA if all are NA
  if(!all(is.na(x))) return(sum(x, na.rm=T))
  if(all(is.na(x))) return(NA)
}

r2_general <-function(preds,actual){ 
  return(1- sum((preds - actual) ^ 2)/sum((actual - mean(actual))^2))
}

RMSE_func <- function(preds, actual){
  return(sqrt(mean((actual - preds)^2)))
}

theme_Publication <- theme_pubr() + theme(
               panel.border =  element_rect(colour="#f0f0f0", fill = NA))+
  theme(plot.title = element_text(hjust = 0.5))


select <- dplyr::select
rename <- dplyr::rename

sf_use_s2(FALSE) #<- this avoids issues with st_intersection
```

#1. load and clean data 
##1.1. ecomon data 
load in the data 
```{r}
stratum <- read_sf("whale_data/EcomonStrata_v4b.shp")
stratum <- st_set_crs(stratum, 4326)
ecomon <- read.csv("ecomon_data/EcoMon_Plankton_Data_v3_7-Data.csv")
ecomon_x <- ecomon[c(1:13)]
ecomon_snames <- colnames(ecomon)
ecomon_snames <- ecomon_snames[grepl("_10m2", ecomon_snames)]

ecomon_y <- ecomon[colnames(ecomon) %in% ecomon_snames]
ecomon_y[is.na(ecomon_y)] <- 0 #NAs should be true 0s

#convert data and create year, month, day columns:
ecomon <- cbind(ecomon_x, ecomon_y)

ecomon$date <- as.Date(ecomon$date, format =  "%d-%b-%y") #b here is the code for month when its words lol
ecomon <- ecomon %>% mutate(day = day(date), month = month(date), year = year(date), week = week(date))
ecomon$haulid <- paste(ecomon$cruise_name, "_", ecomon$station, "_", ecomon$depth)





#now sample grid ID for ecomon table and whale table 
ecomon_spat <- ecomon


coordinates(ecomon_spat) <- ~ lon + lat
ecomon_spat <- st_as_sf(ecomon_spat)

ecomon_spat <- st_set_crs(ecomon_spat, 4326)

#clip to stratum
ecomon_spat <- st_intersection(ecomon_spat,stratum)

#ecomon_extract <- st_join(ecomon_spat, stratum, join = st_nearest_feature)

#checkit 
ggplot()  + geom_sf(data = world) +geom_sf(data = ecomon_spat, aes(colour = as.factor(Region))) +
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)
```

##1.2 Stomach content  
```{r}
load("~/Documents/whales/Whales/ammo21.rdata")
stom <- ammo21 %>% filter(!analcat %in% c("EMPTY", "(Other)", "OTHFIS", "AR"))
#EUPFAM, CLUFAM, CTENOP, COPEPO, AMPHIP, CNIDAR, ENGFAM, CUMACE

stom$haulid <- paste(formatC(stom$cruise6, width=6, flag=0), formatC(stom$station, width=3, flag=0), formatC(stom$stratum, width=4, flag=0), sep='-')

stom$declon <- stom$declon*-1

stom <- stom %>% dplyr::select(haulid, analcat, analsci, collsci, declat, declon, month, day, year, pyamtw, pynam, pdscinam) %>% drop_na()


#now sample grid ID for ecomon table and whale table 
stom_spat <- stom


coordinates(stom_spat) <- ~ declon + declat
stom_spat <- st_as_sf(stom_spat)

stom_spat <- st_set_crs(stom_spat, 4326)

stom_spat <- st_intersection(stom_spat,stratum)
#stom_extract <- st_join(stom_spat, stratum, join = st_nearest_feature)

ggplot()  + geom_sf(data = world)  + geom_sf(data = stom_spat, aes(colour = as.factor(Region))) +
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)
```

##1.3 Trawls
```{r}
load("NEFSC_BTS_2021_all_seasons.RData")
load("stratum.area.RData")
fish_table <- survey$survdat
#this is monthly cpue (wtcpue)

# sum different sexes of same spp together
#okay so now we know that the biomass column was given to them as the calibrated cpue
fish_table <- fish_table %>% 
  group_by(YEAR, SEASON, EST_TOWDATE, LAT, LON, CRUISE6, DEPTH, STATION, STRATUM, SVSPP, COMNAME) %>% 
  dplyr::summarise(wtcpue = sum(BIOMASS), abd = sum(ABUNDANCE), BOTTEMP = mean(BOTTEMP), BOTSALIN = mean(BOTSALIN), SURFTEMP = mean(SURFTEMP), SURFSALIN = mean(SURFSALIN)) 

neus_strata <- stratum.area %>% 
  dplyr::select(STRATUM, STRATUM_AREA) %>% 
  mutate(STRATUM = as.integer(STRATUM)) %>%
  distinct()

fish_table <-  left_join(fish_table, neus_strata, by = c("STRATUM" = "STRATUM"))

fish_table <- filter(fish_table, !is.na(STRATUM_AREA))

fish_table <- fish_table %>%
  mutate(
    # Create a unique haulid
    haulid = paste(formatC(CRUISE6, width=6, flag=0), formatC(STATION, width=3, flag=0), formatC(STRATUM, width=4, flag=0), sep='-'),  
    # Calculate stratum area where needed (use convex hull approach)
    # convert square nautical miles to square kilometers
    STRATUM_AREA = STRATUM_AREA * 3.429904) %>% 
  dplyr::rename(year = YEAR,
         spp = COMNAME,
         date = EST_TOWDATE,
         lat = LAT, 
         lon = LON, 
         depth = DEPTH,
         stratum = STRATUM, 
         stratumarea = STRATUM_AREA) %>%
  filter(
    # remove unidentified spp and non-species
    spp != "" | !is.na(spp), 
    !grepl("EGG", spp), 
    !grepl("UNIDENTIFIED", spp)) %>%
  group_by(haulid, stratum, stratumarea, year, date, lat, lon, depth, spp, SEASON) %>% 
  dplyr::summarise(wtcpue = sumna(wtcpue), BT = mean(BOTTEMP), BSAL = mean(BOTSALIN), SST = mean(SURFTEMP), SSAL = mean(SURFSALIN)) %>% 
  # add temporary region column (this will be replaced with seasonal name)
  mutate(region = "Northeast US") %>% 
  dplyr::select(region, haulid, year, date, lat, lon, stratum, stratumarea, depth, spp, wtcpue, SEASON, BT, SST, BSAL, SSAL) %>% 
  ungroup()


fish_table <- fish_table %>% 
  drop_na(lat,lon)

fish_table$month <- month(as.POSIXlt(fish_table$date))


fish_table_wide <- pivot_wider(fish_table, names_from = spp, values_from = wtcpue)
 #with more than 20% NA

fish_table <- fish_table_wide

#convert NA to 0 
fish_table[, 15:ncol(fish_table)][is.na(fish_table[, 15:ncol(fish_table)])] <- 0

colnames(fish_table) <- gsub(pattern = ' ', replacement = ".", x = colnames(fish_table))
fish_table <- fish_table %>% dplyr::select(-SEASON)


fish_table <- fish_table %>% 
  drop_na(lat,lon)


fish_spat <- fish_table


coordinates(fish_spat) <- ~ lon + lat
fish_spat <- st_as_sf(fish_spat)

fish_spat <- st_set_crs(fish_spat, 4326)


#clip to grid
fish_spat <- st_intersection(fish_spat,stratum) #the stratum are consistently sampled, so I think I need to calculate a stratum level mean centroid 
fish_spat <- fish_spat %>% mutate(gridID = Name)

```

##1.4 whales
```{r}
whale <- read.table("whale_data/baleen_segments.txt", header = TRUE)
whale <- whale %>% dplyr::filter(ModelID == 28) #humpback 

whale$date <- as.Date(whale$StartTime, format = "%m/%d/%Y")
whale <- whale %>% mutate(day = day(date), month = month(date), year = year(date), week = week(date))


#now sample grid ID for whale table and whale table 
whale_survey <- readxl::read_xlsx("whale_data/survey_data.xlsx")
whale_big <- left_join(whale, whale_survey, by = c("SurveyID" = "ID"))

whale_spat <- whale_big

coordinates(whale_spat) <- ~ X_mid + Y_mid
whale_spat <- st_as_sf(whale_spat)

whale_spat <- st_set_crs(whale_spat, 4326)

ggplot()  + geom_sf(data = world)  + geom_sf(data = whale_spat, aes(colour = as.factor(Provider)), size = .2, alpha = .5) +
    coord_sf(xlim=c(-80, -60), ylim=c(35,50), expand = TRUE)+
    theme(legend.position="bottom")+ guides(colour = guide_legend(override.aes = list(size=4), nrow = 2))

whale_spat <- whale

coordinates(whale_spat) <- ~ X_mid + Y_mid
whale_spat <- st_as_sf(whale_spat)

whale_spat <- st_set_crs(whale_spat, 4326)

whale_spat <- st_intersection(whale_spat,stratum)
#whale_extract <- st_join(whale_spat, stratum, join = st_nearest_feature)

ggplot()  + geom_sf(data = world)  + geom_sf(data = whale_spat, aes(colour = as.factor(Region))) +
    coord_sf(xlim=c(-80, -60), ylim=c(35,50), expand = TRUE)+
    theme(legend.position="bottom")
```


#2. Index of abunace - seasonal
##2.1 whales 
Here year will start in December
segment area is in m2. 

check for consistent sampling 
```{r}
#
#whale
whale_samp <- whale_spat %>% st_drop_geometry() %>% drop_na() %>% mutate(season = if_else(month == 12 | month == 1 | month == 2, "winter", if_else(month == 3 | month == 4| month == 5, "spring", if_else(month == 6| month == 7| month == 8, "summer", "fall")))) %>% mutate(year_2 = year-(min(year-1))) %>% mutate()

whale_samp <- whale_samp %>% mutate(year_2 = if_else(month == 12, year_2-1, year_2))

whale_samp %>% ggplot() + geom_histogram(aes(x = year, fill = as.factor(season))) + facet_wrap(~Region, scales = "free_y")


```

plot abundances - p value label 
```{r}

#whale
whale_mon <- whale_spat %>% st_drop_geometry() %>% drop_na() %>% mutate(season = if_else(month == 12 | month == 1 | month == 2, "winter", if_else(month == 3 | month == 4| month == 5, "spring", if_else(month == 6| month == 7| month == 8, "summer", "fall")))) %>% mutate(year_2 = year-(min(year-1))) %>% mutate()

whale_mon <- whale_mon %>% mutate(year_2 = if_else(month == 12, year_2-1, year_2))

whale_mon <- whale_mon %>% group_by(Region, year_2, season) %>% summarise(Individu_1 = sum(Individu_1), length = sum(SegmentAre)) 

whale_mon <- whale_mon %>% drop_na() %>% mutate(AbdPerArea = Individu_1/length) 


#relevel factors for plotting 
whale_mon$Region <- factor(whale_mon$Region, levels = c("GB", "GOM", "SNE", "MAB"))
whale_mon$season <- factor(whale_mon$season, levels = c("winter", "spring", "summer", "fall"))

ggplot(data = whale_mon, aes(x = year_2+1995, y = AbdPerArea*100000)) + geom_point() + geom_smooth(method = "glm", colour = "blue") + labs(x = "year", y = expression("humpback denisty as individuals/100 km" ^2)) + facet_grid(season ~ Region, scales = "free_y") +theme_Publication + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  stat_poly_eq(mapping = aes(
                 color = ifelse(after_stat(p.value) < 0.05, "red", "black"),
                 label = after_stat(round(p.value, 3))))+
  scale_color_identity() 


ggsave(filename = "figures/manuscript/all_abundance.png", plot=last_plot(), width = 25, height=10, units=c("cm"), dpi=500)

```
##2.2 prey 
indices of abundance for stomach content ecomon, trawl data - added categories to summarize at Julia's level 

stomach
```{r}
#stomach - total species specific prey weight/total predator stomachs
  #need to add in a column for unique predators 

stom_mon <- stom_spat %>% st_drop_geometry() %>% drop_na() %>% st_drop_geometry() %>% drop_na() %>% mutate(season = if_else(month == 12 | month == 1 | month == 2, "winter", if_else(month == 3 | month == 4| month == 5, "spring", if_else(month == 6| month == 7| month == 8, "summer", "fall")))) %>% mutate(year_2 = year-(min(year-1))) %>% mutate()

stom_mon <- stom_mon %>% mutate(year_2 = if_else(month == 12, year_2-1, year_2))

stom_mon$category <- paste(stom_mon$analsci, "stomach", sep = "_")

stom_mon <- stom_mon %>%mutate(pdcount = 1) %>% group_by(Region, year_2, season, category) %>% summarise(pyamtw = sum(pyamtw)/sum(pdcount)) %>% ungroup()

  #get rid of empty strings
stom_mon <- stom_mon[!apply(stom_mon, 1, function(x) any(x=="")),] 

stom_mon <- stom_mon %>% ungroup()%>%  pivot_wider(names_from = category, values_from = pyamtw)

stom_mon <- stom_mon[ , colSums(is.na(stom_mon)) < 200]

```

ecomon
```{r}
#ecomon - average abundance per 10m2 per month 
ecomon_mon <- ecomon_spat %>%st_drop_geometry() %>% drop_na() 

ecomon_mon <- ecomon_mon %>% pivot_longer(cols = ctyp_10m2:lopame_10m2, names_to = "spp", values_to = "count_10m2")

ecomon_mon <- ecomon_mon  %>% st_drop_geometry() %>% drop_na() %>% mutate(season = if_else(month == 12 | month == 1 | month == 2, "winter", if_else(month == 3 | month == 4| month == 5, "spring", if_else(month == 6| month == 7| month == 8, "summer", "fall")))) %>% mutate(year_2 = year-(min(year-1))) %>% mutate()

ecomon_mon <- ecomon_mon %>% mutate(year_2 = if_else(month == 12, year_2-1, year_2))

ecomon_types <- read.csv("ecomon_snames.csv")



ecomon_mon <- left_join(ecomon_mon, ecomon_types, by = c("spp" = "spp_code"))

ecomon_mon$category <- paste(ecomon_mon$additional_category, "ecomon", sep = "_")


ecomon_env <- ecomon_mon %>% group_by(Region, year_2, season) %>% summarise(sfc_temp = mean(sfc_temp), sfc_salt = mean(sfc_salt), btm_temp = mean(btm_temp), btm_salt = mean(btm_salt))
                                                                               
ecomon_mon <- ecomon_mon %>% group_by(Region, year_2, category, season) %>% summarise(count_10m2 = mean(count_10m2))

ecomon_mon <- ecomon_mon %>% ungroup()%>%  pivot_wider(names_from = category, values_from = count_10m2)


ecomon_mon <- ecomon_mon[ , colSums((ecomon_mon == 0)) < 200]

ecomon_mon <- left_join(ecomon_mon, ecomon_env, by = c("Region", "year_2", "season"))
```

trawls
```{r}


#trawls 
trawl_mon <- fish_spat %>% st_drop_geometry() %>% drop_na()
trawl_mon <- trawl_mon %>% pivot_longer(cols = FOURSPOT.FLOUNDER:PELAGIC.STINGRAY, names_to = "spp", values_to = "biomass")


fish_types <- read.csv("spp_use.csv")

trawl_mon <- left_join(trawl_mon, fish_types, by = c("spp"))
trawl_mon$other_category <- gsub(pattern = ' ', replacement = "_", x = trawl_mon$other_category)

trawl_mon$category <- paste(trawl_mon$other_category, "trawl", sep = "_")

trawl_mon <- trawl_mon  %>% st_drop_geometry() %>% drop_na() %>% mutate(season = if_else(month == 12 | month == 1 | month == 2, "winter", if_else(month == 3 | month == 4| month == 5, "spring", if_else(month == 6| month == 7| month == 8, "summer", "fall")))) %>% mutate(year_2 = year-(min(year-1))) %>% mutate()


trawl_mon <- trawl_mon %>%  select(year_2, Region, season, biomass, category) %>% group_by(Region, year_2,season, category) %>% summarise(biomass = mean(biomass))

trawl_mon <- trawl_mon %>% pivot_wider(names_from = category, values_from = biomass)

trawl_mon <- trawl_mon[ , colSums((trawl_mon == 0)) < 200]


```

##2.3 together 
relate those indices together and understand what is explaining changes in humback abundance
```{r}
sum <- left_join(whale_mon, stom_mon, by = c("Region" = "Region", "year_2" ="year_2", "season" = "season"))

sum <- left_join(sum, ecomon_mon, by = c("Region" = "Region", "year_2" ="year_2", "season" = "season"))

sum <- left_join(sum, trawl_mon, by = c("Region" = "Region", "year_2" ="year_2", "season" = "season"))

```

regional + seasonal

maybe start by only looking at ones that are changing their distributions? 
```{r}
sum$region_season <- paste(sum$Region, sum$season, sep = "_")
region_seasons <- unique(sum$region_season)
```

###2.3.1 GLM
```{r}
x <- sum %>% select(AMPHIPODA_stomach:mesopelagic_fish_trawl) %>% ungroup()
xnames <- colnames(x)
xnames <- c(xnames, "year_2")
res_glm <- data.frame(matrix(ncol=5, nrow=1))
colnames(res_glm) <- c("names", "deviance", "p_val", "region", "coef")

res_glm_mid <- data.frame(matrix(ncol=5, nrow=1))
colnames(res_glm_mid) <- c("names", "deviance", "p_val", "region", "coef")




  for(b in 1:length(region_seasons)){
    region <- region_seasons[b]
    use <- sum %>% filter(region_season %in% region)
  for (i in 1:length(xnames)) { #for each species in spec_list
  tryCatch({ fit <- glm(as.formula(paste("AbdPerArea ~", xnames[i])), data=use)
  p_val <- summary(fit)$coefficients[2,4]
  dev <- 1 - (summary(fit)$deviance/summary(fit)$null.deviance)
  res_glm_mid[i,1] <- xnames[i]
  res_glm_mid[i,2] <- dev
  res_glm_mid[i,3] <- p_val
  res_glm_mid[i, 4] <- region
  res_glm_mid[i, 5] <- fit$coefficients[2]
}, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }
   res_glm <- rbind(res_glm_mid, res_glm)
  }

res_glm <- res_glm[complete.cases(res_glm),]
res_glm$direction <- ifelse(res_glm$coef > 0, "positive", "negative")

res_glm <- unique(res_glm)
res_glm %>%
  mutate(species_fact = names,
         species = reorder_within(names, deviance, region)) %>%
  ggplot(aes(deviance, species, fill = region)) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~region, scales = "free_y")

#this gets out the top X
res_glm <- res_glm %>% separate(region, into = c("region", "season"))
res_glm$regionseason <- paste(res_glm$region, res_glm$season, sep = "_")

res_glm %>%  group_by(`regionseason`) %>% 
  mutate(
    `regionseason` = as.factor(`regionseason`),
    `names` = reorder_within(`names`,`deviance`,`regionseason`)
  ) %>% 
  slice_max(`deviance`, n =5)%>%
  ggplot(aes(deviance, names, fill = direction)) +
  geom_col(show.legend = TRUE) +
  scale_y_reordered() + ggh4x::facet_grid2(season ~ region, scales = "free_y", independent = "y")  + scale_fill_manual(values = c("light blue", "orange")) + theme_Publication


```
what is the average deviance explained of prey vs. the environment? 
```{r}
res_glm$cat <- ifelse(res_glm$names %in% c("sfc_temp", "sfc_salt", "btm_temp", "btm_salt"), "env", "prey")

res_glm %>% filter(names != "year")%>% ggplot(aes(y = deviance, x = cat, colour = cat)) + geom_boxplot() + stat_compare_means(method = "wilcox")


```

#TO do: 
get environmental variables that are consistent 
should I only use season + region + year combinations that were consistently sampled for each prey item? 
or combine the prey items so we have some consistent version of abundance across the different trawls. 

#3. Distribution shifts
1. create a grid 
2. use only grid cells that have consistent sampling over time 
3. calculate grid cell of density for each year (season)
4. relate that to environmental and prey species. 

##3.1 Whales
###3.1 1. create grid 
```{r}

globe_bb <- matrix(c(-84,  20,
                      -84,  50,
                      -52, 50,
                     -52, 20,
                     -84,  20), byrow = TRUE, ncol = 2) %>%
  list() %>% 
  st_polygon() %>% 
  st_sfc(., crs = 4326)

# generate grid of 128 x 128 tiles (each .5 x .5 degrees)
grid <- st_make_grid(globe_bb, n = c(126, 126), 
                                 crs = 4326, what = 'polygons') %>%
  st_sf('geometry' = ., data.frame('ID' = 1:length(.)))



whale_spat <- whale


coordinates(whale_spat) <- ~ X_mid + Y_mid
whale_spat <- st_as_sf(whale_spat)

whale_spat <- st_set_crs(whale_spat, 4326)



#checkit
ggplot(data = world) + geom_sf() +geom_sf(data = globe_bb) + geom_point(data = whale, aes(x = X_mid, y = Y_mid)) + 
    coord_sf(xlim=c(-90, -20), ylim=c(20,60), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

ggplot(data = world) + geom_sf() +geom_sf(data = grid) + geom_point(data = whale, aes(x = X_mid, y = Y_mid)) + 
    coord_sf(xlim=c(-90, -20), ylim=c(20,60), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))



whale_spat <- st_intersection(whale_spat,grid)

ggplot(data = world) + geom_sf()+ geom_sf(data = whale_spat, aes(colour = ID.1)) + 
    coord_sf(xlim=c(-90, -20), ylim=c(20,60), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))
```


###3.1.2 consistent grid 
find grid cells that were consistently sampled through time 

```{r}
whale_spat <- whale_spat %>% mutate(gridID = ID.1)


#annual
#unique years 
length(unique(whale_spat$year))

grouped <- whale_spat %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

whale_spat <- left_join(whale_spat, grouped, by = "gridID")

whale_coords <- whale_spat %>% st_coordinates() 

whale_spat <- cbind(whale_spat, whale_coords)

whale_spat <- whale_spat %>% filter(Y > 40) #this just gets out the gulf of maine/new england area 

grouped <- whale_spat %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

#50 grid cells have 18 years worth of data so lets start with those 

ggplot() +geom_sf(data = grid) + geom_sf(data = subset(whale_spat, whale_spat$n > 17), aes(colour = n)) + geom_sf(data = world)+ 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))
```

50 grid cells have 18 years worth of data so lets start with those 

make sure that there wasn't a shift in timing of surveys 

```{r}
whale_use <- whale_spat %>% filter(n > 17)
seasonal_check <- whale_spat %>% filter(gridID %in% whale_use$gridID)
seasonal_check %>% ggplot(aes(x = month, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~year)
```
###3.1.3 Spring centroids 
Look at only the spring as not to bias towards differential sampling 

```{r}

whale_use <- whale_spat %>% filter(n > 17)
whale_use <- whale_use %>% filter(month == 4 | month == 5 | month == 6 )

ggplot(data = world) + geom_sf()+ geom_sf(data = whale_use, aes(colour = month), size = 2, shape = 15)+
    coord_sf(xlim=c(-75, -65), ylim=c(40,44), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black")) + facet_wrap(~year)



###1.3.2 calculate it 
grids <- st_centroid(grid) 

centroids <- st_coordinates(grids)
centroids <- cbind(grids, centroids)
centroids <- centroids %>% rename(X_mid = X, Y_mid = Y) %>% st_drop_geometry()

#group by grid cell and year to find gridded abundances 
whale_use <- whale_use %>% st_drop_geometry() %>% group_by(gridID, year)  %>% summarise(Individu_1 = sum(Individu_1), length = sum(SegmentAre)) 

whale_use <- whale_use %>% drop_na() %>% group_by(gridID, year) %>% summarise(AbdPerArea = Individu_1/length) 

whale_use <- left_join(whale_use, centroids, by = c("gridID"= "ID"))

#find mean centroids of abundance (this is gridded so a bit coarse

#for loop for each year 
years <- unique(whale_use$year)
weighted_centroids <- data.frame(matrix(nrow =1, ncol = 3))
colnames(weighted_centroids) <- c("X", "Y", "year")

for (i in 1:length(years)){
  year_use = years[i]
  use <- whale_use %>% filter(year %in% year_use)
  whale_mid <- use 
  coordinates(whale_mid) <- ~ X_mid + Y_mid
  whale_cent <- wt.centroid(whale_mid, p = "AbdPerArea")
  weighted_centroids[i, 1] <- whale_cent@coords[1]
  weighted_centroids[i, 2] <- whale_cent@coords[2]
  weighted_centroids[i, 3] <- year_use
  
  
}

#distance from boston 
boston <- c(-71.021856, 42.399123)
weighted_centroids_dist <- distGeo(weighted_centroids, boston)

weighted_centroids$boston <- weighted_centroids_dist

weighted_centroids_hump <- weighted_centroids

#distance from stellwagen 
stellwagen <- st_read("stellb/data/NationalMarineSanctuary.shp")
stellwagen_2 <- st_centroid(stellwagen)

coordinates(weighted_centroids) <- ~ X + Y
weighted_centroids <- st_as_sf(weighted_centroids)

weighted_centroids <- st_set_crs(weighted_centroids, 4326)
stellwagen <- st_set_crs(stellwagen, 4326)
weighted_centroids_dist <- st_distance(weighted_centroids, stellwagen)

weighted_centroids$stellwagen <- weighted_centroids_dist

stellwagen_2 <- st_set_crs(stellwagen_2, 4326)
weighted_centroids_dist <- st_distance(weighted_centroids, stellwagen_2)

weighted_centroids$stellwagen_center <- weighted_centroids_dist



#get back into dataframe 
cent_coords <- weighted_centroids %>% st_coordinates() 

weighted_centroids <- cbind(weighted_centroids, cent_coords)
weighted_centroids <- weighted_centroids %>% st_drop_geometry()
weighted_centroids$stellwagen <- as.numeric(weighted_centroids$stellwagen)
weighted_centroids$stellwagen_center <- as.numeric(weighted_centroids$stellwagen_center)
weighted_centroids_hump <- weighted_centroids



ggplot(data = weighted_centroids_hump,aes(x = year, y = Y)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = year, y = Y), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point() + labs(y = "latitude")

ggplot(data = weighted_centroids_hump,aes(x = year, y = boston)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = year, y = boston), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point() + labs(y = "distance to boston")

ggplot(data = weighted_centroids_hump,aes(x = year, y = stellwagen)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = year, y = stellwagen), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point() + labs(y = "distance to stellwagen")




ggplot(data = world) + geom_sf()+ geom_point(data = subset(whale_use, whale_use$AbdPerArea > 0), aes(x = X_mid, y = Y_mid, colour = AbdPerArea), size = 2, shape = 15) +scale_colour_gradient2(
  low = "#56B1F7",
  mid = "yellow",
  high = "dark red",
  midpoint = .00000001,
  aesthetics = "colour"
)+
    coord_sf(xlim=c(-75, -65), ylim=c(40,44), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black")) + facet_wrap(~year)



weighted_centroids_hump$heatwave <- ifelse(weighted_centroids_hump$year < 2013, "pre_2012_heatwave", "post_2012_heatwave")

ggplot(data = world) + geom_sf()+ geom_point(data = weighted_centroids_hump, aes(x = X, y = Y, colour = heatwave)) + 
    coord_sf(xlim=c(-75, -65), ylim=c(40,45), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black")) + theme(legend.position = "bottom")

weighted_centroids_hump %>% ggplot(aes(y = stellwagen_center, x = heatwave, colour = heatwave)) + geom_boxplot() + stat_compare_means(method = "wilcox")

weighted_centroids_hump <- weighted_centroids_hump %>% select(-boston, -stellwagen)


p1 <- ggplot(data = world) + geom_sf()+ geom_point(data =  weighted_centroids_hump, aes(x = X, y = Y, colour = year)) + geom_sf(data = stellwagen_2, colour = "red") + geom_sf(data=stellwagen, fill = NA, colour = "red")+
    coord_sf(xlim=c(-75, -65), ylim=c(40,45), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black")) + theme(legend.position=c(.8,.3))

p1

p2 <- ggplot(data = weighted_centroids_hump,aes(x = year, y = stellwagen_center)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = year, y = stellwagen_center), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point() + labs(y = "distance to stellwagen")

p3 <- p1 + p2+ plot_annotation(tag_levels = 'A')
p3
```
###3.1.3 Area occupied 
- has their total area occupied changed? (grid cells they were present in/area surveyed)

```{r}
occ <- whale_use %>% filter(AbdPerArea > 0) %>% group_by(year) %>% tally()
occ$total_grid <- length(unique(whale_use$gridID))
occ <- occ %>% mutate(perc_occupied = n/total_grid)

ggplot(data = occ ,aes(x = year, y = perc_occupied)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()

```
They might be expanding their range a bit 


looks like maybe they are moving inshore - can relate this shift to changes in prey abundances. 


###3.1.4	Using areas differently? 
-	What proportion of their density is in each area (stellwagen vs. southeast channel) in each year 

```{r}

stellwagen <- st_read("stellb/data/NationalMarineSanctuary.shp")
GSC_pots <- st_read("Great_South_Channel_Restricted_Trap-Pot_Area/Great_South_Channel_Restricted_Trap-Pot_Area.shp")
GSC_dredge <- st_read("HabitatClamDredgeExemption_GreatSouthChannel2020616/HabitatClamDredgeExemption_GreatSouthChannel2020.shp")

ggplot() + geom_sf(data = world)+ geom_sf(data = stellwagen, aes(colour = "stellwagen"), fill = NA)+ geom_sf(data = GSC_pots,aes(colour = "GSC Pots"), fill = NA) + geom_sf(data = GSC_dredge, aes(colour = "GSC dredge"), fill = NA) + 
    coord_sf(xlim=c(-75, -65), ylim=c(40,45), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))
```
what was their density in each area per year (number of individuals/area occupied) - I am not sure if I then need to divide by the area of each protected area 

```{r}
whale_use <- whale_spat %>% filter(n > 17)
whale_use <- whale_use %>% filter(month == 4 | month == 5 | month == 6 )


GSC_pots <- st_transform(GSC_pots, 4326)
st_crs(GSC_pots)
st_crs(whale_use)
GSC_area <- st_intersection(GSC_pots, whale_use) #these are all the whale poinst within GSC 

stellwagen <- st_transform(stellwagen, 4326)
st_crs(stellwagen)
stellwagen_area <- st_intersection(stellwagen, whale_use) #these are all the whale poinst within GSC 



ggplot() + geom_sf(data = world)+ geom_sf(data = GSC_area) + geom_sf(data = stellwagen_area) +
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

stellwagen_use <- stellwagen_area %>% st_drop_geometry() %>% group_by(year) %>% summarise(stell_densityarea = (sum(Individu_1)/sum(SegmentAre))/AREA_KM, stell_density = sum(Individu_1)/sum(SegmentAre)) %>% distinct()

GSC_area1 <- GSC_pots %>% mutate(area = set_units(st_area(GSC_pots), "km2"))
GSC_area1 <- GSC_area1$area
GSC_area1 <- GSC_area1 %>% drop_units

#I 

stellwagen_area1 <- stellwagen %>% mutate(area = set_units(st_area(stellwagen), "km2")) #nice it is the same 

GSC_use <- GSC_area %>% st_drop_geometry() %>% group_by(year) %>% summarise(GSC_densityarea = (sum(Individu_1)/sum(SegmentAre))/GSC_area1, GSC_density = sum(Individu_1)/sum(SegmentAre))

changing_area <- left_join(stellwagen_use, GSC_use, by = "year")
changing_area <- changing_area %>% pivot_longer(stell_densityarea:GSC_density, values_to = "density", names_to = "area")
changing_area <- changing_area %>% separate(area, into = c("protected_area", "metric"))

ggplot(data = changing_area) + geom_line(aes(x = year, y = density, colour = protected_area)) + facet_wrap(~metric, scales = "free_y")

#I think it is okay to just look at density (because area surveyed should take into account spatial differences)

changing_area %>% filter(metric == "density") %>% ggplot(aes(x = year, y = density, colour = protected_area)) + geom_line() +
  stat_poly_line() +
  stat_poly_eq()
```
To me it looks like how they area using Stellwagen bank has changed, while the way they are using GSC has not over the years - r2 is the same if you do it both/area or not 

##3.2 prey shifts
relate those shifts to prey 

Grouped 
prey grouped by order/julias groupings  
###3.2.1 ecomon

```{r}
#now sample grid ID for ecomon table and whale table 
ecomon_spat <- ecomon


coordinates(ecomon_spat) <- ~ lon + lat
ecomon_spat <- st_as_sf(ecomon_spat)

ecomon_spat <- st_set_crs(ecomon_spat, 4326)


#clip to grid
ecomon_spat <- st_intersection(ecomon_spat,stratum) #the stratum are consistently sampled, so I think I need to calculate a stratum level mean centroid 
ecomon_spat <- ecomon_spat %>% mutate(gridID = Name)

ecomon_coords <- ecomon_spat %>% st_coordinates() 

ecomon_spat <- cbind(ecomon_spat, ecomon_coords)

whale_use1 <- whale_spat %>% filter(n > 17)

ecomon_use <- ecomon_spat %>% filter(Region == "GOM" & year >= min(whale_use$year) & month %in% c(4,5,6))



ggplot(data = world) + geom_sf()+ geom_sf(data = ecomon_use, aes(colour = as.factor(gridID))) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


```


```{r}
grids <- st_centroid(stratum) 

centroids <- st_coordinates(grids)
centroids <- cbind(grids, centroids)
centroids <- centroids %>% rename(X_mid = X, Y_mid = Y) %>% st_drop_geometry()

#group by grid cell and year to find gridded abundances 
ecomon_long <- ecomon_use %>% pivot_longer(cols = ecomon_snames, names_to = "spp", values_to = "abd")

#get out only common ones 
spp_use <- ecomon_long %>% st_drop_geometry() %>% filter(abd > 0) %>% group_by(spp) %>% tally()
write.csv(spp_use, "ecomon_use_nmes.csv")

ecomon_types <- read.csv("ecomon_snames.csv")

ecomon_long <- left_join(ecomon_long, ecomon_types, by = c("spp" = "spp_code"))

ecomon_long$category <- paste(ecomon_long$additional_category, "ecomon", sep = "_")

ecomon_use <- ecomon_long %>% st_drop_geometry() %>% group_by(gridID, year, spp)  %>% summarise(abd = mean(abd)) %>% drop_na()

#group by julia category 
ecomon_use <- ecomon_long %>% st_drop_geometry() %>% group_by(gridID, year, category)  %>% summarise(abd = mean(abd)) %>% drop_na()

ecomon_use <- left_join(ecomon_use, centroids, by = c("gridID"= "Name"))


ecomon_grouped <- ecomon_use %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

ecomon_use <- left_join(ecomon_use, ecomon_grouped, by = "gridID")

ggplot(data = world) + geom_sf()+ geom_point(data = subset(ecomon_use, ecomon_use$n >16), aes(colour = n, x = X_mid, y = Y_mid)) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

ecomon_use <- ecomon_use %>% filter(n==17)

ecomon_use <- ecomon_use[!apply(ecomon_use, 1, function(x) any(x=="")),] 

#find mean centroids of abundance (this is gridded by stratum so a bit coarse

#for loop for each year 
years <- unique(ecomon_use$year)
specs <- unique(ecomon_use$spp)
specs <- unique(ecomon_use$category)
weighted_centroids <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids) <- c("X", "Y", "year", "spec")

weighted_centroids_mid <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids_mid) <- c("X", "Y", "year", "spec")


for (a in 1:length(years)){
  year_use <- years[a]
  use <- ecomon_use %>% filter(year %in% year_use)
  for (i in 1:length(specs)) {tryCatch({ 
  specs_use = specs[i]
  use2 <- use %>% filter(category %in% specs_use)
  ecomon_mid <- use2 
  coordinates(ecomon_mid) <- ~ X_mid + Y_mid
  ecomon_cent <- wt.centroid(ecomon_mid, p = "abd")
  weighted_centroids_mid[i, 1] <- ecomon_cent@coords[1]
  weighted_centroids_mid[i, 2] <- ecomon_cent@coords[2]
  weighted_centroids_mid[i, 3] <- year_use
  weighted_centroids_mid[i, 4] <- specs[i]
  
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
  weighted_centroids <- rbind(weighted_centroids_mid, weighted_centroids)
}

#distance from stellwagen center  
stellwagen_center <- c(-70.32125, 42.4044)
weighted_centroids_dist <- distGeo(weighted_centroids, stellwagen_center)

weighted_centroids$stellwagen_center <- weighted_centroids_dist

weighted_centroids_ecomon <- weighted_centroids


```

It looks like there is something going on here with prey centroids and humpback distributions 

###3.2.2 trawls



```{r}
load("NEFSC_BTS_2021_all_seasons.RData")
load("stratum.area.RData")
fish_table <- survey$survdat
#this is monthly cpue (wtcpue)

# sum different sexes of same spp together
#okay so now we know that the biomass column was given to them as the calibrated cpue
fish_table <- fish_table %>% 
  group_by(YEAR, SEASON, EST_TOWDATE, LAT, LON, CRUISE6, DEPTH, STATION, STRATUM, SVSPP, COMNAME) %>% 
  dplyr::summarise(wtcpue = sum(BIOMASS), abd = sum(ABUNDANCE), BOTTEMP = mean(BOTTEMP), BOTSALIN = mean(BOTSALIN), SURFTEMP = mean(SURFTEMP), SURFSALIN = mean(SURFSALIN)) 

neus_strata <- stratum.area %>% 
  dplyr::select(STRATUM, STRATUM_AREA) %>% 
  mutate(STRATUM = as.integer(STRATUM)) %>%
  distinct()

fish_table <-  left_join(fish_table, neus_strata, by = c("STRATUM" = "STRATUM"))

fish_table <- filter(fish_table, !is.na(STRATUM_AREA))

fish_table <- fish_table %>%
  mutate(
    # Create a unique haulid
    haulid = paste(formatC(CRUISE6, width=6, flag=0), formatC(STATION, width=3, flag=0), formatC(STRATUM, width=4, flag=0), sep='-'),  
    # Calculate stratum area where needed (use convex hull approach)
    # convert square nautical miles to square kilometers
    STRATUM_AREA = STRATUM_AREA * 3.429904) %>% 
  dplyr::rename(year = YEAR,
         spp = COMNAME,
         date = EST_TOWDATE,
         lat = LAT, 
         lon = LON, 
         depth = DEPTH,
         stratum = STRATUM, 
         stratumarea = STRATUM_AREA) %>%
  filter(
    # remove unidentified spp and non-species
    spp != "" | !is.na(spp), 
    !grepl("EGG", spp), 
    !grepl("UNIDENTIFIED", spp)) %>%
  group_by(haulid, stratum, stratumarea, year, date, lat, lon, depth, spp, SEASON) %>% 
  dplyr::summarise(wtcpue = sumna(wtcpue), BT = mean(BOTTEMP), BSAL = mean(BOTSALIN), SST = mean(SURFTEMP), SSAL = mean(SURFSALIN)) %>% 
  # add temporary region column (this will be replaced with seasonal name)
  mutate(region = "Northeast US") %>% 
  dplyr::select(region, haulid, year, date, lat, lon, stratum, stratumarea, depth, spp, wtcpue, SEASON, BT, SST, BSAL, SSAL) %>% 
  ungroup()


fish_table <- fish_table %>% 
  drop_na(lat,lon)

fish_table$month <- month(as.POSIXlt(fish_table$date))


fish_table_wide <- pivot_wider(fish_table, names_from = spp, values_from = wtcpue)
 #with more than 20% NA

fish_table <- fish_table_wide

#convert NA to 0 
fish_table[, 15:ncol(fish_table)][is.na(fish_table[, 15:ncol(fish_table)])] <- 0

colnames(fish_table) <- gsub(pattern = ' ', replacement = ".", x = colnames(fish_table))
fish_table <- fish_table %>% dplyr::select(-SEASON)


fish_table <- fish_table %>% 
  drop_na(lat,lon)


fish_spat <- fish_table


coordinates(fish_spat) <- ~ lon + lat
fish_spat <- st_as_sf(fish_spat)

fish_spat <- st_set_crs(fish_spat, 4326)


#clip to grid
fish_spat <- st_intersection(fish_spat,stratum) #the stratum are consistently sampled, so I think I need to calculate a stratum level mean centroid 
fish_spat <- fish_spat %>% mutate(gridID = Name)

fish_coords <- fish_spat %>% st_coordinates() 

fish_spat <- cbind(fish_spat, fish_coords)


fish_spat <- fish_spat %>% filter(Region == "GOM" & year >= min(whale_use$year) & month %in% c(4,5,6)) #this just gets out the same area as the whale data and same months



ggplot(data = world) + geom_sf()+ geom_sf(data = fish_spat, aes(colour = as.factor(gridID))) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


fish_use <- fish_spat

```

```{r}
grids <- st_centroid(stratum) 

centroids <- st_coordinates(grids)
centroids <- cbind(grids, centroids)
centroids <- centroids %>% rename(X_mid = X, Y_mid = Y) %>% st_drop_geometry()

#group by grid cell and year to find gridded abundances 
fish_long <- fish_use %>% pivot_longer(cols = c(FOURSPOT.FLOUNDER:PELAGIC.STINGRAY), names_to = "spp", values_to = "abd")

#get out only common ones 
spp_use <- fish_long %>% st_drop_geometry() %>% filter(abd > 0) %>% group_by(spp) %>% tally()

fish_types <- read.csv("spp_use.csv")

fish_long <- left_join(fish_long, fish_types, by = c("spp"))

fish_use <- fish_long %>% st_drop_geometry() %>% group_by(gridID, year, spp)  %>% summarise(abd = mean(abd)) %>% drop_na()

#group by category 
fish_long$category <- paste(fish_long$other_category, "trawls", sep = "_")

fish_use <- fish_long %>% st_drop_geometry() %>% group_by(gridID, year, category)  %>% summarise(abd = mean(abd)) %>% drop_na()

fish_use <- left_join(fish_use, centroids, by = c("gridID"= "Name"))


fish_grouped <- fish_use %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

fish_use <- left_join(fish_use, fish_grouped, by = "gridID")

ggplot(data = world) + geom_sf()+ geom_point(data = subset(fish_use, fish_use$n >16), aes(colour = n, x = X_mid, y = Y_mid)) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

fish_use <- fish_use %>% filter(n==17)

fish_use <- fish_use[!apply(fish_use, 1, function(x) any(x=="")),] 

#find mean centroids of abundance (this is gridded by stratum so a bit coarse

#for loop for each year 
years <- unique(fish_use$year)
specs <- unique(fish_use$spp)
specs <- unique(fish_use$category)
weighted_centroids <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids) <- c("X", "Y", "year", "spec")

weighted_centroids_mid <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids_mid) <- c("X", "Y", "year", "spec")


for (a in 1:length(years)){
  year_use <- years[a]
  use <- fish_use %>% filter(year %in% year_use)
  for (i in 1:length(specs)) {tryCatch({ 
  specs_use = specs[i]
  use2 <- use %>% filter(category %in% specs_use)
  fish_mid <- use2 
  coordinates(fish_mid) <- ~ X_mid + Y_mid
  fish_cent <- wt.centroid(fish_mid, p = "abd")
  weighted_centroids_mid[i, 1] <- fish_cent@coords[1]
  weighted_centroids_mid[i, 2] <- fish_cent@coords[2]
  weighted_centroids_mid[i, 3] <- year_use
  weighted_centroids_mid[i, 4] <- specs[i]
  
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
  weighted_centroids <- rbind(weighted_centroids_mid, weighted_centroids)
}

#distance from stellwagen center  
stellwagen_center <- c(-70.32125, 42.4044)
weighted_centroids_dist <- distGeo(weighted_centroids, stellwagen_center)

weighted_centroids$stellwagen_center <- weighted_centroids_dist

weighted_centroids_fish <- weighted_centroids

weighted_centroids_fish <- weighted_centroids_fish %>% drop_na(spec)
```

###3.2.3 stomach

```{r}
#now sample grid ID for ecomon table and whale table 
stom_spat <- stom


coordinates(stom_spat) <- ~ declon + declat
stom_spat <- st_as_sf(stom_spat)

stom_spat <- st_set_crs(stom_spat, 4326)

stom_spat <- st_intersection(stom_spat,stratum)

stom_spat <- stom_spat %>% mutate(gridID = Name)

stom_coords <- stom_spat %>% st_coordinates() 

stom_spat <- cbind(stom_spat, stom_coords)

stom_use <- stom_spat  %>% filter(Region == "GOM" & year >= min(whale_use$year) & month %in% c(4,5,6)) #this just gets out the same



ggplot(data = world) + geom_sf()+ geom_sf(data = stom_use, aes(colour = gridID)) + 
    coord_sf(xlim=c(-90, -20), ylim=c(20,60), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


#Make sure they are consistently sampled in all of the same 17 years 
grouped <- stom_use %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

stom_use <- left_join(stom_use, grouped, by = "gridID")

ggplot() +geom_sf(data = grid) + geom_sf(data = subset(stom_use, stom_use$n > 17), aes(colour = n)) + geom_sf(data = world)+ 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

ggplot() +geom_sf(data = grid) + geom_sf(data = stom_use, aes(colour = n)) + geom_sf(data = world)+ 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


```
Lets try it with the stomach content, same years overlapping region (not grid cells, just min and max lat/long)

```{r}
grids <- st_centroid(stratum) 

centroids <- st_coordinates(grids)
centroids <- cbind(grids, centroids)
centroids <- centroids %>% rename(X_mid = X, Y_mid = Y) %>% st_drop_geometry()

stom_use <- stom_use[!apply(stom_use, 1, function(x) any(x=="")),] #remove empty strings

stom_use <- stom_use %>% st_drop_geometry() %>% dplyr::select(pyamtw, haulid, analsci, day, month, year, Region, gridID) %>% group_by(haulid, analsci, day, month, year, Region, gridID) %>% summarise(pyamtw = sum(pyamtw)) 
#add in the zeros 

stome_spread <- stom_use %>% pivot_wider(names_from = "analsci", values_from = "pyamtw")

stome_spread[is.na(stome_spread)] <- 0

stom_use2 <- stome_spread %>% pivot_longer(AMMODYTIDAE:CYCLOPTERIDAE, names_to = "analsci", values_to = "pyamtw")

stom_use2$category <- paste(stom_use2$analsci, "stomach", sep = "_")

#group by grid cell and year to find gridded abundances 

stom_use2 <- stom_use2 %>% group_by(gridID, year, category)  %>% summarise(pyamtw = mean(pyamtw)) %>% drop_na()

stom_use2 <- left_join(stom_use2, centroids, by = c("gridID"= "Name"))



#find mean centroids of abundance (this is gridded so a bit coarse

#for loop for each year 
years <- unique(stom_use2$year)
specs <- unique(stom_use2$category)
weighted_centroids <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids) <- c("X", "Y", "year", "spec")

weighted_centroids_mid <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids_mid) <- c("X", "Y", "year", "spec")


for (a in 1:length(years)){
  year_use <- years[a]
  use <- stom_use2 %>% filter(year %in% year_use)
  for (i in 1:length(specs)) {tryCatch({ 
  specs_use = specs[i]
  use2 <- use %>% filter(category %in% specs_use)
  stom_mid <- use2 
  coordinates(stom_mid) <- ~ X_mid + Y_mid
  stom_cent <- wt.centroid(stom_mid, p = "pyamtw")
  weighted_centroids_mid[i, 1] <- stom_cent@coords[1]
  weighted_centroids_mid[i, 2] <- stom_cent@coords[2]
  weighted_centroids_mid[i, 3] <- year_use
  weighted_centroids_mid[i, 4] <- specs[i]
  
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
  weighted_centroids <- rbind(weighted_centroids_mid, weighted_centroids)
}

#distance from boston 
weighted_centroids_dist <- distGeo(weighted_centroids, stellwagen_center)

weighted_centroids$stellwagen_center <- weighted_centroids_dist

weighted_centroids_stom <- weighted_centroids


ggplot(data = world) + geom_sf()+ geom_point(data = subset(weighted_centroids_stom, weighted_centroids_stom$year > 2000), aes(x = X, y = Y, colour = year)) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black")) 

```
###3.2.4 Env 
Average env variables 
add in OISST temperature 

Lets just take from the trawl data because I know that has  a lot already in it 
```{r}
xdata <- ecomon

```

```{r}
#take gulf of maine seasonal temps (overlapping months with the whales)

xdata <- xdata %>% filter(month %in% c(4, 5, 6)) 
xdata_spat <- xdata 

coordinates(xdata_spat) <- ~ lon + lat
xdata_spat <- st_as_sf(xdata_spat)

xdata_spat <- st_set_crs(xdata_spat, 4326)

xdata_spat <- st_intersection(xdata_spat,stratum)

xdata_coords <- xdata_spat %>% st_coordinates() 

xdata_spat <- cbind(xdata_spat, xdata_coords)


ggplot(data = world) + geom_sf()+ geom_sf(data = xdata_spat, aes(colour = Region), size = .2) + 
    coord_sf(xlim=c(-90, -60), ylim=c(35,55), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


use_years <- whale_use %>% ungroup() %>% select(year) %>% unique()
use_years <- as.vector(use_years$year)

#xdata_use <- xdata_spat %>% filter(est_year >= min(whale_use$year))
xdata_use <- xdata_spat %>% filter(year %in% use_years & Region == "GOM")


```
take the average annual temp and salinity of those grid cells 
```{r}
xdata_sum <- xdata_use %>% st_drop_geometry() %>% group_by(year) %>%  summarise_at(c("sfc_temp", "sfc_salt", "btm_temp", "btm_salt"), mean, na.rm = TRUE)
```


##3.3 prey glms 
on distance
loop through and figure out if significant relationship between humpback and other species mean centroids 
```{r}
weighted_centroids_hump$spec <- "hump"
weighted_centroids_ecomon$heatwave <-   ifelse(weighted_centroids_ecomon$year < 2013, "pre_2012_heatwave", "post_2012_heatwave")

weighted_centroids_stom$heatwave <-   ifelse(weighted_centroids_stom$year < 2013, "pre_2012_heatwave", "post_2012_heatwave")

weighted_centroids_fish$heatwave <-   ifelse(weighted_centroids_fish$year < 2013, "pre_2012_heatwave", "post_2012_heatwave")



centroids_both <- rbind(weighted_centroids_hump, weighted_centroids_ecomon, weighted_centroids_stom, weighted_centroids_fish)

centroids_both <- centroids_both %>% unique()


centroids_wide <- centroids_both %>% select(-X, -Y) %>% pivot_wider(names_from = spec, values_from = stellwagen_center) 

centroids_wide <- centroids_wide %>%  mutate(across(where(is.character), ~na_if(., "NULL")))


#remove columns with a bunch of nas
centroids_wide <- centroids_wide %>% select(which(colMeans(is.na(.)) < 0.2))

centroids_wide <- left_join(centroids_wide, xdata_sum, by = c("year"))

centroids_wide <- centroids_wide %>% drop_na(hump)
centroids_wide_group <- centroids_wide
```

make sure I am using an okay distribution 

```{r}

fit <- glm(hump ~ year, data = centroids_wide)
plot(fit)

#looks like the residuals are pretty normally distributed
```



```{r}

centroid_glm <- data.frame(matrix(ncol=4, nrow=1))
colnames(centroid_glm) <- c("names", "deviance", "pvalue", "coef")

xnames <- colnames(centroids_wide)
xnames <- xnames[xnames != "hump"]

for (i in 1:length(xnames)) { #for each species in spec_list
  tryCatch({ fit <- glm(as.formula(paste("hump ~", xnames[i])), data=centroids_wide)
  p <- summary(fit)$coefficients[2,4] 
  dev <- 1 - (summary(fit)$deviance/summary(fit)$null.deviance)
  centroid_glm[i,1] <- xnames[i]
  centroid_glm[i,2] <- dev
  centroid_glm[i,3] <- p
  centroid_glm[i, 4] <- fit$coefficients[2]
}, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}




centroid_glm$direction <- ifelse(centroid_glm$coef > 0, "positive", "negative")

centroid_glm %>% top_n(deviance,n = 15) %>% 
  mutate(names = fct_reorder(names, deviance)) %>%
  ggplot(aes(x=names, y=deviance, fill = direction)) +
    geom_bar(stat="identity", width=.4) +
    coord_flip() + labs(y = "deviance ",x = element_blank())


nam <- centroid_glm %>% top_n(deviance,n = 15) 
names <- nam$names

for (i in 1:length(names)) {
  dat <- centroids_wide %>% select(hump, names[i])
  colnames(dat) <- c("hump", "x")
p <- ggplot(data = dat ,aes(x = x, y = hump)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = x, y = hump), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()+ labs(title = names[i])

print(p)
}

```

```{r}
centroid_glm$cat <- ifelse(centroid_glm$names %in% c("sfc_temp", "sfc_salt", "btm_temp", "btm_salt"), "env", "prey")
```


```{r}
centroid_glm %>% filter(names != "year") %>% ggplot(aes(y = deviance, x = cat, colour = cat)) + geom_boxplot() + stat_compare_means(method = "wilcox")
```

##3.4 timing 
Timing of peak abundance in GOM 
only use consistently sampled spatial area - find mean_ct of their abundance - is it earlier or later? 
```{r}
whale <- read.table("whale_data/baleen_segments.txt", header = TRUE)
whale <- whale %>% dplyr::filter(ModelID == 28) #humpback 

whale$date <- as.Date(whale$StartTime, format = "%m/%d/%Y")
whale <- whale %>% mutate(day = day(date), month = month(date), year = year(date), week = week(date))


#now sample grid ID for whale table and whale table 
whale_spat <- whale


coordinates(whale_spat) <- ~ X_mid + Y_mid
whale_spat <- st_as_sf(whale_spat)

whale_spat <- st_set_crs(whale_spat, 4326)

whale_spat <- st_intersection(whale_spat,stratum)
#whale_extract <- st_join(whale_spat, stratum, join = st_nearest_feature)

ggplot()  + geom_sf(data = world)  + geom_sf(data = whale_spat, aes(colour = as.factor(Region))) +
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)
```

```{r}
whale_gom <- whale_spat %>% st_drop_geometry() %>% filter(Region == "GOM")

whale_gom %>% ggplot(aes(x = month, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)

#whale_gom <- whale_gom %>% filter(month == 4 | month == 5 & year != 2013)

whale_gom %>% ggplot(aes(x = month, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)


#what is their peak timing of surveyed abundance each year?
whale_gom$jd <- yday(whale_gom$date)
whale_gom$week <- week(whale_gom$date)
whale_gom$week2 <- ifelse(whale_gom$week == 15 | whale_gom$week == 16, 15, ifelse(whale_gom$week == 17 | whale_gom$week == 18, 17, ifelse(whale_gom$week == 19 | whale_gom$week == 20, 19, ifelse(whale_gom$week == 21 | whale_gom$week == 22, 21, NA))))

whale_gom$AbdPerArea <- whale_gom$Individu_1/whale_gom$SegmentAre

#only get out days that were sampled every year 
weeks_use <- whale_gom %>% select(year, week2, AbdPerArea)  %>% group_by(year,week2) %>% summarise(AbdPerArea = mean(AbdPerArea))
weeks_use <- weeks_use %>% pivot_wider(names_from = "week2", values_from = AbdPerArea)




#only use those consistently sampled 2-week periods 
whale_gom <- whale_gom %>% ungroup() %>% drop_na(week2) %>% filter(!year %in% c(1999, 2002, 2010, 2011, 2012, 2013))

whale_gom %>% ggplot(aes(x = week2, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)

whale_ct <- whale_gom %>% mutate(mx = week2*AbdPerArea)

whale_ct <- whale_ct %>% ungroup %>% group_by(year) %>% mutate(ct = (sum(mx))/(sum(AbdPerArea)), avg_abd = mean(AbdPerArea), sum_abd = sum(AbdPerArea))

whale_ct <- whale_ct %>% select(year, ct, avg_abd, sum_abd) %>% unique()


#whale_ct$ct_day <- whale_ct$ct - 31 - 29 - 31

whale_ct %>% ggplot(aes(x = year, y = ct)) + geom_point() + labs(y = "Central tendency of abundance (two-week period)", main = "Gulf of Maine")+ geom_smooth(aes(x = year, y = ct), method = "loess", colour = "blue")
```
#4. Figures

make one big plot of the gulf of maine 

##4.1 abundance 
```{r}
abd1 <- ggplot(data = subset(whale_mon, whale_mon$Region == "GOM" & whale_mon$season == "spring"), aes(x = year_2+1995, y = AbdPerArea*100000)) + geom_point() + geom_smooth(method = "glm", colour = "blue") + labs(x = "year", y = expression("humpback denisty as individuals/100 km" ^2)) +theme_Publication + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  stat_poly_eq(use_label(c("p", "adj.R2"))) 

abd2 <- res_glm %>%  filter(regionseason == "GOM_spring") %>% 
  slice_max(`deviance`, n =10) %>%
  mutate(`names` = reorder_within(`names`,`deviance`,`regionseason`)
  ) %>% 
  ggplot(aes(deviance, names, fill = direction)) +
  geom_col(show.legend = TRUE) +
  scale_y_reordered() + scale_fill_manual(values = c("light blue", "orange")) + theme_Publication

abd1
abd2
```


##4.2 distribution 
```{r}
densa <- ggplot(data = world) + geom_sf()+ geom_point(data =  weighted_centroids_hump, aes(x = X, y = Y, colour = year)) + geom_sf(data=stellwagen, fill = NA, colour = "red")+ geom_sf(data = GSC_pots,colour = "blue", fill = NA) + geom_sf(data = GSC_dredge, colour = "black", fill = NA) +
    coord_sf(xlim=c(-75, -65), ylim=c(40,45), expand = TRUE) + theme_Publication+ theme(panel.background = element_rect(fill = "white", colour = "black")) + theme(legend.position=c(.8,.3)) +
  scale_colour_gradient(low = "yellow", high = "red", na.value = NA) +
  geom_sf_text(data = stellwagen, aes(label = "Stellwagen Bank"), size = 3, colour = "red",
    nudge_x = 0.4, nudge_y = 0.52)+
  geom_sf_text(data = GSC_pots, aes(label = "Great South Channel"), size = 3, colour = "blue",
    nudge_x = 0.1, nudge_y = -0.7)






densb <- ggplot(data = weighted_centroids_hump,aes(x = year, y = stellwagen_center)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = year, y = stellwagen_center), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point() + theme_Publication + labs(y = "distance to stellwagen")

densc <- changing_area %>% filter(metric == "density") %>% ggplot(aes(x = year, y = density*100000, colour = protected_area)) + geom_line() +
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point() + theme_Publication + labs(y = expression("humpback denisty as individuals/100 km" ^2))+ scale_colour_manual(values = c("blue", "red"), labels = c("Great South Channel", "Stellwagen Bank")) + theme(legend.position=c(.52,.91))+
  labs(color=NULL) 
densc

densd <- centroid_glm %>% top_n(deviance,n = 15) %>% 
  mutate(names = fct_reorder(names, deviance)) %>%
  ggplot(aes(x=names, y=deviance, fill = direction)) +
    geom_bar(stat="identity", width=.4) +
    coord_flip() + labs(y = "deviance ",x = element_blank())+ theme_Publication + scale_fill_manual(values = c("light blue", "orange")) + theme(legend.position=c(.8,.3)) 


densa
densb
densc
densd



areaa <- ggplot(data = occ ,aes(x = year, y = perc_occupied)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point() + theme_Publication + labs(y = "% of surveyed occupied")
areaa



areab <- whale_ct %>% ggplot(aes(x = year, y = ct)) + geom_point() + labs(y = "Central tendency of abundance (two-week period)", main = "Gulf of Maine")+ geom_smooth(aes(x = year, y = ct), method = "glm", colour = "blue") +  stat_poly_eq(use_label(c("p", "R2"))) + theme_Publication
time1



p_abd <- abd1 + abd2 + plot_annotation(tag_levels = 'A')
p_abd

ggsave(filename = "figures/manuscript/abundance.png", plot=last_plot(), width = 25, height=15, units=c("cm"), dpi=500)

p_dist <- (densa + densb)/ (densc + densd) + plot_annotation(tag_levels = 'A')
p_dist

ggsave(filename = "figures/manuscript/distribution.png", plot=last_plot(), width = 35, height=25, units=c("cm"), dpi=500)


p_non <- areaa + areab
p_non

ggsave(filename = "figures/manuscript/timing_area.png", plot=last_plot(), width = 25, height=15, units=c("cm"), dpi=500)
```


#__________
##species level 

###ECOMON
```{r}
#now sample grid ID for ecomon table and whale table 
ecomon_spat <- ecomon


coordinates(ecomon_spat) <- ~ lon + lat
ecomon_spat <- st_as_sf(ecomon_spat)

ecomon_spat <- st_set_crs(ecomon_spat, 4326)


#clip to grid
ecomon_spat <- st_intersection(ecomon_spat,stratum) #the stratum are consistently sampled, so I think I need to calculate a stratum level mean centroid 
ecomon_spat <- ecomon_spat %>% mutate(gridID = Name)

ecomon_coords <- ecomon_spat %>% st_coordinates() 

ecomon_spat <- cbind(ecomon_spat, ecomon_coords)

whale_use1 <- whale_spat %>% filter(n > 17)

ecomon_use <- ecomon_spat %>% filter(Region == "GOM" & year >= min(whale_use$year) & month %in% c(4,5,6))



ggplot(data = world) + geom_sf()+ geom_sf(data = ecomon_use, aes(colour = as.factor(gridID))) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


```


```{r}
grids <- st_centroid(stratum) 

centroids <- st_coordinates(grids)
centroids <- cbind(grids, centroids)
centroids <- centroids %>% rename(X_mid = X, Y_mid = Y) %>% st_drop_geometry()

#group by grid cell and year to find gridded abundances 
ecomon_long <- ecomon_use %>% pivot_longer(cols = ecomon_snames, names_to = "spp", values_to = "abd")

#get out only common ones 
spp_use <- ecomon_long %>% st_drop_geometry() %>% filter(abd > 0) %>% group_by(spp) %>% tally()
write.csv(spp_use, "ecomon_use_nmes.csv")

ecomon_types <- read.csv("ecomon_snames.csv")

ecomon_long <- left_join(ecomon_long, ecomon_types, by = c("spp" = "spp_code"))

ecomon_long$category <- paste(ecomon_long$sname, "ecomon", sep = "_")

ecomon_use <- ecomon_long %>% st_drop_geometry() %>% group_by(gridID, year, spp)  %>% summarise(abd = mean(abd)) %>% drop_na()

#group by julia category 
ecomon_use <- ecomon_long %>% st_drop_geometry() %>% group_by(gridID, year, category)  %>% summarise(abd = mean(abd)) %>% drop_na()

ecomon_use <- left_join(ecomon_use, centroids, by = c("gridID"= "Name"))


ecomon_grouped <- ecomon_use %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

ecomon_use <- left_join(ecomon_use, ecomon_grouped, by = "gridID")

ggplot(data = world) + geom_sf()+ geom_point(data = subset(ecomon_use, ecomon_use$n >16), aes(colour = n, x = X_mid, y = Y_mid)) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

ecomon_use <- ecomon_use %>% filter(n==17)

ecomon_use <- ecomon_use[!apply(ecomon_use, 1, function(x) any(x=="")),] 

#find mean centroids of abundance (this is gridded by stratum so a bit coarse

#for loop for each year 
years <- unique(ecomon_use$year)
specs <- unique(ecomon_use$spp)
specs <- unique(ecomon_use$category)
weighted_centroids <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids) <- c("X", "Y", "year", "spec")

weighted_centroids_mid <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids_mid) <- c("X", "Y", "year", "spec")


for (a in 1:length(years)){
  year_use <- years[a]
  use <- ecomon_use %>% filter(year %in% year_use)
  for (i in 1:length(specs)) {tryCatch({ 
  specs_use = specs[i]
  use2 <- use %>% filter(category %in% specs_use)
  ecomon_mid <- use2 
  coordinates(ecomon_mid) <- ~ X_mid + Y_mid
  ecomon_cent <- wt.centroid(ecomon_mid, p = "abd")
  weighted_centroids_mid[i, 1] <- ecomon_cent@coords[1]
  weighted_centroids_mid[i, 2] <- ecomon_cent@coords[2]
  weighted_centroids_mid[i, 3] <- year_use
  weighted_centroids_mid[i, 4] <- specs[i]
  
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
  weighted_centroids <- rbind(weighted_centroids_mid, weighted_centroids)
}

#distance from stellwagen center  
stellwagen_center <- c(-70.32125, 42.4044)
weighted_centroids_dist <- distGeo(weighted_centroids, stellwagen_center)

weighted_centroids$stellwagen_center <- weighted_centroids_dist

weighted_centroids_ecomon <- weighted_centroids


```

It looks like there is something going on here with prey centroids and humpback distributions 

###TRAWLS



```{r}
load("NEFSC_BTS_2021_all_seasons.RData")
load("stratum.area.RData")
fish_table <- survey$survdat
#this is monthly cpue (wtcpue)

# sum different sexes of same spp together
#okay so now we know that the biomass column was given to them as the calibrated cpue
fish_table <- fish_table %>% 
  group_by(YEAR, SEASON, EST_TOWDATE, LAT, LON, CRUISE6, DEPTH, STATION, STRATUM, SVSPP, COMNAME) %>% 
  dplyr::summarise(wtcpue = sum(BIOMASS), abd = sum(ABUNDANCE), BOTTEMP = mean(BOTTEMP), BOTSALIN = mean(BOTSALIN), SURFTEMP = mean(SURFTEMP), SURFSALIN = mean(SURFSALIN)) 

neus_strata <- stratum.area %>% 
  dplyr::select(STRATUM, STRATUM_AREA) %>% 
  mutate(STRATUM = as.integer(STRATUM)) %>%
  distinct()

fish_table <-  left_join(fish_table, neus_strata, by = c("STRATUM" = "STRATUM"))

fish_table <- filter(fish_table, !is.na(STRATUM_AREA))

fish_table <- fish_table %>%
  mutate(
    # Create a unique haulid
    haulid = paste(formatC(CRUISE6, width=6, flag=0), formatC(STATION, width=3, flag=0), formatC(STRATUM, width=4, flag=0), sep='-'),  
    # Calculate stratum area where needed (use convex hull approach)
    # convert square nautical miles to square kilometers
    STRATUM_AREA = STRATUM_AREA * 3.429904) %>% 
  dplyr::rename(year = YEAR,
         spp = COMNAME,
         date = EST_TOWDATE,
         lat = LAT, 
         lon = LON, 
         depth = DEPTH,
         stratum = STRATUM, 
         stratumarea = STRATUM_AREA) %>%
  filter(
    # remove unidentified spp and non-species
    spp != "" | !is.na(spp), 
    !grepl("EGG", spp), 
    !grepl("UNIDENTIFIED", spp)) %>%
  group_by(haulid, stratum, stratumarea, year, date, lat, lon, depth, spp, SEASON) %>% 
  dplyr::summarise(wtcpue = sumna(wtcpue), BT = mean(BOTTEMP), BSAL = mean(BOTSALIN), SST = mean(SURFTEMP), SSAL = mean(SURFSALIN)) %>% 
  # add temporary region column (this will be replaced with seasonal name)
  mutate(region = "Northeast US") %>% 
  dplyr::select(region, haulid, year, date, lat, lon, stratum, stratumarea, depth, spp, wtcpue, SEASON, BT, SST, BSAL, SSAL) %>% 
  ungroup()


fish_table <- fish_table %>% 
  drop_na(lat,lon)

fish_table$month <- month(as.POSIXlt(fish_table$date))


fish_table_wide <- pivot_wider(fish_table, names_from = spp, values_from = wtcpue)
 #with more than 20% NA

fish_table <- fish_table_wide

#convert NA to 0 
fish_table[, 15:ncol(fish_table)][is.na(fish_table[, 15:ncol(fish_table)])] <- 0

colnames(fish_table) <- gsub(pattern = ' ', replacement = ".", x = colnames(fish_table))
fish_table <- fish_table %>% dplyr::select(-SEASON)


fish_table <- fish_table %>% 
  drop_na(lat,lon)


fish_spat <- fish_table


coordinates(fish_spat) <- ~ lon + lat
fish_spat <- st_as_sf(fish_spat)

fish_spat <- st_set_crs(fish_spat, 4326)


#clip to grid
fish_spat <- st_intersection(fish_spat,stratum) #the stratum are consistently sampled, so I think I need to calculate a stratum level mean centroid 
fish_spat <- fish_spat %>% mutate(gridID = Name)

fish_coords <- fish_spat %>% st_coordinates() 

fish_spat <- cbind(fish_spat, fish_coords)


fish_spat <- fish_spat %>% filter(Region == "GOM" & year >= min(whale_use$year) & month %in% c(4,5,6)) #this just gets out the same area as the whale data and same months



ggplot(data = world) + geom_sf()+ geom_sf(data = fish_spat, aes(colour = as.factor(gridID))) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


fish_use <- fish_spat

```

```{r}
grids <- st_centroid(stratum) 

centroids <- st_coordinates(grids)
centroids <- cbind(grids, centroids)
centroids <- centroids %>% rename(X_mid = X, Y_mid = Y) %>% st_drop_geometry()

#group by grid cell and year to find gridded abundances 
fish_long <- fish_use %>% pivot_longer(cols = c(FOURSPOT.FLOUNDER:PELAGIC.STINGRAY), names_to = "spp", values_to = "abd")

#get out only common ones 
spp_use <- fish_long %>% st_drop_geometry() %>% filter(abd > 0) %>% group_by(spp) %>% tally()

fish_types <- read.csv("spp_by_year_selected.csv")
fish_types$COMNAME <- gsub(pattern = ' ', replacement = ".", x = fish_types$COMNAME)

fish_long <- left_join(fish_long, fish_types, by = c("spp" = "COMNAME"))

fish_use <- fish_long %>% st_drop_geometry() %>% group_by(gridID, year, spp)  %>% summarise(abd = mean(abd)) %>% drop_na()

#group by category 
fish_long$category <- paste(fish_long$spp, "trawls", sep = "_")

fish_use <- fish_long %>% st_drop_geometry() %>% group_by(gridID, year, category)  %>% summarise(abd = mean(abd)) %>% drop_na()

fish_use <- left_join(fish_use, centroids, by = c("gridID"= "Name"))


fish_grouped <- fish_use %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

fish_use <- left_join(fish_use, fish_grouped, by = "gridID")

ggplot(data = world) + geom_sf()+ geom_point(data = subset(fish_use, fish_use$n >16), aes(colour = n, x = X_mid, y = Y_mid)) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

fish_use <- fish_use %>% filter(n==17)

fish_use <- fish_use[!apply(fish_use, 1, function(x) any(x=="")),] 

#find mean centroids of abundance (this is gridded by stratum so a bit coarse

#for loop for each year 
years <- unique(fish_use$year)
specs <- unique(fish_use$spp)
specs <- unique(fish_use$category)
weighted_centroids <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids) <- c("X", "Y", "year", "spec")

weighted_centroids_mid <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids_mid) <- c("X", "Y", "year", "spec")


for (a in 1:length(years)){
  year_use <- years[a]
  use <- fish_use %>% filter(year %in% year_use)
  for (i in 1:length(specs)) {tryCatch({ 
  specs_use = specs[i]
  use2 <- use %>% filter(category %in% specs_use)
  fish_mid <- use2 
  coordinates(fish_mid) <- ~ X_mid + Y_mid
  fish_cent <- wt.centroid(fish_mid, p = "abd")
  weighted_centroids_mid[i, 1] <- fish_cent@coords[1]
  weighted_centroids_mid[i, 2] <- fish_cent@coords[2]
  weighted_centroids_mid[i, 3] <- year_use
  weighted_centroids_mid[i, 4] <- specs[i]
  
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
  weighted_centroids <- rbind(weighted_centroids_mid, weighted_centroids)
}

#distance from stellwagen center  
stellwagen_center <- c(-70.32125, 42.4044)
weighted_centroids_dist <- distGeo(weighted_centroids, stellwagen_center)

weighted_centroids$stellwagen_center <- weighted_centroids_dist

weighted_centroids_fish <- weighted_centroids

weighted_centroids_fish <- weighted_centroids_fish %>% drop_na(spec)
```

###STOMACH

```{r}
#now sample grid ID for ecomon table and whale table 
stom_spat <- stom


coordinates(stom_spat) <- ~ declon + declat
stom_spat <- st_as_sf(stom_spat)

stom_spat <- st_set_crs(stom_spat, 4326)

stom_spat <- st_intersection(stom_spat,stratum)

stom_spat <- stom_spat %>% mutate(gridID = Name)

stom_coords <- stom_spat %>% st_coordinates() 

stom_spat <- cbind(stom_spat, stom_coords)

stom_use <- stom_spat  %>% filter(Region == "GOM" & year >= min(whale_use$year) & month %in% c(4,5,6)) #this just gets out the same



ggplot(data = world) + geom_sf()+ geom_sf(data = stom_use, aes(colour = gridID)) + 
    coord_sf(xlim=c(-90, -20), ylim=c(20,60), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


#Make sure they are consistently sampled in all of the same 17 years 
grouped <- stom_use %>% st_drop_geometry() %>% select(year, gridID) %>% unique %>% group_by(gridID) %>% tally() 

stom_use <- left_join(stom_use, grouped, by = "gridID")

ggplot() +geom_sf(data = grid) + geom_sf(data = subset(stom_use, stom_use$n > 17), aes(colour = n)) + geom_sf(data = world)+ 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))

ggplot() +geom_sf(data = grid) + geom_sf(data = stom_use, aes(colour = n)) + geom_sf(data = world)+ 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


```
Lets try it with the stomach content, same years overlapping region (not grid cells, just min and max lat/long)

```{r}
grids <- st_centroid(stratum) 

centroids <- st_coordinates(grids)
centroids <- cbind(grids, centroids)
centroids <- centroids %>% rename(X_mid = X, Y_mid = Y) %>% st_drop_geometry()

stom_use <- stom_use[!apply(stom_use, 1, function(x) any(x=="")),] #remove empty strings

stom_use <- stom_use %>% st_drop_geometry() %>% dplyr::select(pyamtw, haulid, pynam, day, month, year, Region, gridID) %>% group_by(haulid, pynam, day, month, year, Region, gridID) %>% summarise(pyamtw = sum(pyamtw)) 
#add in the zeros 
stom_use$pynam <- gsub(pattern = ' ', replacement = "_", stom_use$pynam)

stome_spread <- stom_use %>% pivot_wider(names_from = "pynam", values_from = "pyamtw")

stome_spread[is.na(stome_spread)] <- 0

stom_use2 <- stome_spread %>% pivot_longer(AMMODYTES_SP:STRING, names_to = "pynam", values_to = "pyamtw")

stom_use2$category <- paste(stom_use2$pynam, "stomach", sep = "_")

#group by grid cell and year to find gridded abundances 

stom_use2 <- stom_use2 %>% group_by(gridID, year, category)  %>% summarise(pyamtw = mean(pyamtw)) %>% drop_na()

stom_use2 <- left_join(stom_use2, centroids, by = c("gridID"= "Name"))



#find mean centroids of abundance (this is gridded so a bit coarse

#for loop for each year 
years <- unique(stom_use2$year)
specs <- unique(stom_use2$category)
weighted_centroids <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids) <- c("X", "Y", "year", "spec")

weighted_centroids_mid <- data.frame(matrix(nrow =1, ncol = 4))
colnames(weighted_centroids_mid) <- c("X", "Y", "year", "spec")


for (a in 1:length(years)){
  year_use <- years[a]
  use <- stom_use2 %>% filter(year %in% year_use)
  for (i in 1:length(specs)) {tryCatch({ 
  specs_use = specs[i]
  use2 <- use %>% filter(category %in% specs_use)
  stom_mid <- use2 
  coordinates(stom_mid) <- ~ X_mid + Y_mid
  stom_cent <- wt.centroid(stom_mid, p = "pyamtw")
  weighted_centroids_mid[i, 1] <- stom_cent@coords[1]
  weighted_centroids_mid[i, 2] <- stom_cent@coords[2]
  weighted_centroids_mid[i, 3] <- year_use
  weighted_centroids_mid[i, 4] <- specs[i]
  
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
  weighted_centroids <- rbind(weighted_centroids_mid, weighted_centroids)
}

#distance from boston 
weighted_centroids_dist <- distGeo(weighted_centroids, stellwagen_center)

weighted_centroids$stellwagen_center <- weighted_centroids_dist

weighted_centroids_stom <- weighted_centroids


ggplot(data = world) + geom_sf()+ geom_point(data = subset(weighted_centroids_stom, weighted_centroids_stom$year > 2000), aes(x = X, y = Y, colour = year)) + 
    coord_sf(xlim=c(-80, -60), ylim=c(40,50), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black")) 

```
###1.3.1 prey glms 
on distance
loop through and figure out if significant relationship between humpback and other species mean centroids 
```{r}
weighted_centroids_hump$spec <- "hump"
weighted_centroids_ecomon$heatwave <-   ifelse(weighted_centroids_ecomon$year < 2013, "pre_2012_heatwave", "post_2012_heatwave")

weighted_centroids_stom$heatwave <-   ifelse(weighted_centroids_stom$year < 2013, "pre_2012_heatwave", "post_2012_heatwave")

weighted_centroids_fish$heatwave <-   ifelse(weighted_centroids_fish$year < 2013, "pre_2012_heatwave", "post_2012_heatwave")



centroids_both <- rbind(weighted_centroids_hump, weighted_centroids_ecomon, weighted_centroids_stom, weighted_centroids_fish)

centroids_both <- centroids_both %>% unique()


centroids_wide <- centroids_both %>% select(-X, -Y) %>% pivot_wider(names_from = spec, values_from = stellwagen_center) 

centroids_wide <- centroids_wide %>%  mutate(across(where(is.character), ~na_if(., "NULL")))


#remove columns with a bunch of nas
centroids_wide <- centroids_wide %>% select(which(colMeans(is.na(.)) < 0.2))

centroids_wide <- left_join(centroids_wide, xdata_sum, by = c("year"))

centroids_wide <- centroids_wide %>% drop_na(hump)
```

make sure I am using an okay distribution 

```{r}

fit <- glm(hump ~ year, data = centroids_wide)
plot(fit)

#looks like the residuals are pretty normally distributed
```



```{r}

centroid_glm <- data.frame(matrix(ncol=4, nrow=1))
colnames(centroid_glm) <- c("names", "deviance", "pvalue", "coef")

xnames <- colnames(centroids_wide)
xnames <- xnames[xnames != "hump"]

for (i in 1:length(xnames)) { #for each species in spec_list
  tryCatch({ fit <- glm(as.formula(paste("hump ~", xnames[i])), data=centroids_wide)
  p <- summary(fit)$coefficients[2,4] 
  dev <- 1 - (summary(fit)$deviance/summary(fit)$null.deviance)
  centroid_glm[i,1] <- xnames[i]
  centroid_glm[i,2] <- dev
  centroid_glm[i,3] <- p
  centroid_glm[i, 4] <- fit$coefficients[2]
}, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}




centroid_glm$direction <- ifelse(centroid_glm$coef > 0, "positive", "negative")

centroid_glm %>% top_n(deviance,n = 15) %>% 
  mutate(names = fct_reorder(names, deviance)) %>%
  ggplot(aes(x=names, y=deviance, fill = direction)) +
    geom_bar(stat="identity", width=.4) +
    coord_flip() + labs(y = "deviance ",x = element_blank())+ scale_fill_manual(values = c("light blue", "orange"))


nam <- centroid_glm %>% top_n(deviance,n = 15) 
names <- nam$names

for (i in 1:length(names)) {
  dat <- centroids_wide %>% select(hump, names[i])
  colnames(dat) <- c("hump", "x")
p <- ggplot(data = dat ,aes(x = x, y = hump)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = x, y = hump), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()+ labs(title = names[i])

print(p)
}

```

```{r}
centroid_glm$cat <- ifelse(centroid_glm$names %in% c("sfc_temp", "sfc_salt", "btm_temp", "btm_salt"), "env", "prey")
```


```{r}
centroid_glm %>% filter(names != "year") %>% ggplot(aes(y = deviance, x = cat, colour = cat)) + geom_boxplot() + stat_compare_means(method = "wilcox")
```

##1.4 Environment 

####1.4.1 NAO AMO 
start with the NAO and AMO 

```{r}
NAO <- read.csv("climatologies/NAO.csv")
AMO <- read.csv("climatologies/AMO_uns.csv")
AtlTri <- read.csv("climatologies/AtlTri.csv")

NAO <- NAO %>%
   mutate(NAO_average = rowMeans(across(X1:X12))) %>% select(Year,NAO_average) 

AMO <- AMO %>%
   mutate(AMO_average = rowMeans(across(X1:X12)))  %>% select(Year,AMO_average) 

AtlTri <- AtlTri %>%
   mutate(AtlTri_average = rowMeans(across(X1:X12))) %>% select(Year,AtlTri_average) 

weighted_centroids_hump <- left_join(weighted_centroids_hump, NAO, by = c("year" = "Year"))

weighted_centroids_hump <- left_join(weighted_centroids_hump, AMO, by = c("year" = "Year"))
weighted_centroids_hump <- left_join(weighted_centroids_hump, AtlTri, by = c("year" = "Year"))


```


```{r}
ggplot(data = weighted_centroids_hump ,aes(x = NAO_average, y = Y)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = NAO_average, y = Y), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


ggplot(data = weighted_centroids_hump ,aes(x = NAO_average, y = boston)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = NAO_average, y = boston), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()

ggplot(data = weighted_centroids_hump ,aes(x = AMO_average, y = Y)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = AMO_average, y = Y), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


ggplot(data = weighted_centroids_hump ,aes(x = AMO_average, y = boston)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = AMO_average, y = boston), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()



#take out earlier years 
ggplot(data = subset(weighted_centroids_hump,weighted_centroids_hump$year > 2002)  ,aes(x = NAO_average, y = Y)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = NAO_average, y = Y), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


ggplot(data = subset(weighted_centroids_hump,weighted_centroids_hump$year > 2002) ,aes(x = NAO_average, y = boston)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = NAO_average, y = boston), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()

ggplot(data = subset(weighted_centroids_hump,weighted_centroids_hump$year > 2002) ,aes(x = AMO_average, y = Y)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = AMO_average, y = Y), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


ggplot(data = subset(weighted_centroids_hump,weighted_centroids_hump$year > 2002) ,aes(x = AMO_average, y = boston)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = AMO_average, y = boston), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()

```

Those aren't really explaining much 

####1.4.2 Average env variables 
add in OISST temperature 

Lets just take from the trawl data because I know that has  a lot already in it 
```{r}
xdata <- ecomon

```

```{r}
#take gulf of maine seasonal temps (overlapping months with the whales)

xdata <- xdata %>% filter(month %in% c(4, 5, 6)) 
xdata_spat <- xdata 

coordinates(xdata_spat) <- ~ lon + lat
xdata_spat <- st_as_sf(xdata_spat)

xdata_spat <- st_set_crs(xdata_spat, 4326)

xdata_spat <- st_intersection(xdata_spat,stratum)

xdata_coords <- xdata_spat %>% st_coordinates() 

xdata_spat <- cbind(xdata_spat, xdata_coords)


ggplot(data = world) + geom_sf()+ geom_sf(data = xdata_spat, aes(colour = Region), size = .2) + 
    coord_sf(xlim=c(-90, -60), ylim=c(35,55), expand = TRUE) + theme(panel.background = element_rect(fill = "white", colour = "black"))


use_years <- whale_use %>% ungroup() %>% select(year) %>% unique()
use_years <- as.vector(use_years$year)

#xdata_use <- xdata_spat %>% filter(est_year >= min(whale_use$year))
xdata_use <- xdata_spat %>% filter(year %in% use_years & Region == "GOM")


```
take the average annual temp and salinity of those grid cells 
```{r}
xdata_sum <- xdata_use %>% st_drop_geometry() %>% group_by(year) %>%  summarise_at(c("sfc_temp", "sfc_salt", "btm_temp", "btm_salt"), mean, na.rm = TRUE)
```

```{r}
weighted_centroids_hump <- left_join(weighted_centroids_hump, xdata_sum, by = c("year" = "year"))


```



make a long dataframe to show all of the env. variable relationships at once 
```{r}
centroids_long <- weighted_centroids_hump %>% pivot_longer(cols = c(NAO_average:btm_salt), names_to = "env_var", values_to = "env_val")

centroids_long %>% ggplot(aes(x = env_val, y = boston)) + geom_point() + facet_wrap(~env_var, scales = "free_x")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()
```



####1.4.3 climate velocity 

```{r}
remotes::install_github("jebyrnes/hadsstR")

library(hadsstr)


```


```{r}
library(maptools)
data(wrld_simpl)
library(raster)

HadiSST <- load_hadsst(file = "HadISST_sst.nc")
rasts <- get_all_rasters(HadiSST, years = 1997:2018)

velocity <- select_raster(rasts, "velocity_magnitude")

e <- extent(-160,10,30,60)
rasts1 <- crop(rasts, extent(-72,-68,41,43))

cuts=c(-50, -25, 0,25,50,100,150,200, 250) #set breaks
pal <- colorRampPalette(c("lightblue","red"))
plot(rasts1$velocity_magnitude, xlim=c(-75,-65), ylim=c(40,45), axes=TRUE, main = "climate velocity (km/year)", breaks=cuts, col = pal(8))
plot(wrld_simpl, add=TRUE)

pal <- colorRampPalette(c("darkblue", "lightblue", "red", "darkred"))
cuts=c(-1,-.5, -.25, -.1, 0,.25, .5,.75, 1) #set breaks
plot(rasts1$linear_change, xlim=c(-75,-65), ylim=c(40,45), axes=TRUE, main = "linear change (deg C)", breaks=cuts, col = pal(9))
plot(wrld_simpl, add=TRUE)



points(subset(dolph_extract_2, dolph_extract_2$AbdPerArea > 0))

ggplot(data = world) + geom_sf() + geom_raster(data = rasts1$velocity_magnitude)+ 
    coord_sf(xlim=c(-80, -72), ylim=c(32,38), expand = TRUE)+ theme(panel.background = element_rect(fill = "white", colour = "black"))
  
  geom_point(data = subset(dolph_extract_2, dolph_extract_2$AbdPerArea > 0), aes(x = coords.x1, y = coords.x2, colour = AbdPerArea))  + facet_wrap(~Year_) +  scale_colour_gradientn(colours = c("yellow", "red", "darkred"))

```


#Timing of peak abundance in GOM 
only use consistently sampled spatial area - find mean_ct of their abundance - is it earlier or later? 
```{r}
whale <- read.table("whale_data/baleen_segments.txt", header = TRUE)
whale <- whale %>% dplyr::filter(ModelID == 28) #humpback 

whale$date <- as.Date(whale$StartTime, format = "%m/%d/%Y")
whale <- whale %>% mutate(day = day(date), month = month(date), year = year(date), week = week(date))


#now sample grid ID for whale table and whale table 
whale_spat <- whale


coordinates(whale_spat) <- ~ X_mid + Y_mid
whale_spat <- st_as_sf(whale_spat)

whale_spat <- st_set_crs(whale_spat, 4326)

whale_spat <- st_intersection(whale_spat,stratum)
#whale_extract <- st_join(whale_spat, stratum, join = st_nearest_feature)

ggplot()  + geom_sf(data = world)  + geom_sf(data = whale_spat, aes(colour = as.factor(Region))) +
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)
```

```{r}
whale_gom <- whale_spat %>% st_drop_geometry() %>% filter(Region == "GOM")

whale_gom %>% ggplot(aes(x = month, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)

#whale_gom <- whale_gom %>% filter(month == 4 | month == 5 & year != 2013)

whale_gom %>% ggplot(aes(x = month, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)


#what is their peak timing of surveyed abundance each year?
whale_gom$jd <- yday(whale_gom$date)
whale_gom$week <- week(whale_gom$date)
whale_gom$week2 <- ifelse(whale_gom$week == 15 | whale_gom$week == 16, 15, ifelse(whale_gom$week == 17 | whale_gom$week == 18, 17, ifelse(whale_gom$week == 19 | whale_gom$week == 20, 19, ifelse(whale_gom$week == 21 | whale_gom$week == 22, 21, NA))))

whale_gom$AbdPerArea <- whale_gom$Individu_1/whale_gom$SegmentAre

#only get out days that were sampled every year 
weeks_use <- whale_gom %>% select(year, week2, AbdPerArea)  %>% group_by(year,week2) %>% summarise(AbdPerArea = mean(AbdPerArea))
weeks_use <- weeks_use %>% pivot_wider(names_from = "week2", values_from = AbdPerArea)




#only use those consistently sampled 2-week periods 
whale_gom <- whale_gom %>% ungroup() %>% drop_na(week2) %>% filter(!year %in% c(1999, 2002, 2010, 2011, 2012, 2013))

whale_gom %>% ggplot(aes(x = week2, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)

whale_ct <- whale_gom %>% mutate(mx = week2*AbdPerArea)

whale_ct <- whale_ct %>% ungroup %>% group_by(year) %>% mutate(ct = (sum(mx))/(sum(AbdPerArea)), avg_abd = mean(AbdPerArea), sum_abd = sum(AbdPerArea))

whale_ct <- whale_ct %>% select(year, ct, avg_abd, sum_abd) %>% unique()


#whale_ct$ct_day <- whale_ct$ct - 31 - 29 - 31

whale_ct %>% ggplot(aes(x = year, y = ct)) + geom_point() + labs(y = "Central tendency of abundance (two-week period)", main = "Gulf of Maine")+ geom_smooth(aes(x = year, y = ct), method = "loess", colour = "blue")
```
is this related to prey at all - what about the environment? 

```{r}
whale_ct <- left_join(whale_ct, NAO, by = c("year" = "Year"))

whale_ct <- left_join(whale_ct, AMO, by = c("year" = "Year"))
whale_ct <- left_join(whale_ct, AtlTri, by = c("year" = "Year"))

```


```{r}
ggplot(data = whale_ct ,aes(x = NAO_average, y = ct_day)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = NAO_average, y = ct_day), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()



ggplot(data = whale_ct ,aes(x = AMO_average, y = ct_day)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = AMO_average, y = ct_day), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


```

```{r}
#now sample grid ID for ecomon table and whale table 
xdata_spat <- xdata


coordinates(xdata_spat) <- ~ decdeg_beglon + decdeg_beglat
xdata_spat <- st_as_sf(xdata_spat)

xdata_spat <- st_set_crs(xdata_spat, 4326)

xdata_spat <- st_intersection(xdata_spat,stratum)

xdata_coords <- xdata_spat %>% st_coordinates() 

xdata_spat <- cbind(xdata_spat, xdata_coords)

#xdata_use <- xdata_spat %>% filter(est_year >= min(whale_use$year))
xdata_use <- xdata_spat %>% filter(Region == "GOM")



```
take the average annual temp and salinity of those grid cells 
```{r}
xdata_sum <- xdata_use %>% st_drop_geometry() %>% group_by(est_year) %>%  summarise_at(c("surftemp", "surfsalin", "bottemp", "botsalin"), mean, na.rm = TRUE)
```

```{r}
whale_ct <- left_join(whale_ct, xdata_sum, by = c("year" = "est_year"))


```

```{r}
ggplot(data = whale_ct ,aes(x = surftemp, y = ct_day)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = surftemp, y = ct_day), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()



ggplot(data = whale_ct ,aes(x = bottemp, y = ct_day)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = bottemp, y = ct_day), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


ggplot(data = whale_ct ,aes(x = surfsalin, y = ct_day)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = surfsalin, y = ct_day), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()



ggplot(data = whale_ct ,aes(x = botsalin, y = ct_day)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = botsalin, y = ct_day), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


```
interesting - the timing is related to salinity which is probably related to prey 

```{r}
#now sample grid ID for whale table and whale table 
stom_spat <- stom


coordinates(stom_spat) <- ~ declon + declat
stom_spat <- st_as_sf(stom_spat)

stom_spat <- st_set_crs(stom_spat, 4326)

stom_spat <- st_intersection(stom_spat,stratum)
#stom_extract <- st_join(stom_spat, stratum, join = st_nearest_feature)

ggplot()  + geom_sf(data = world)  + geom_sf(data = stom_spat, aes(colour = as.factor(Region))) +
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)
```

```{r}
stom_gom <- stom_spat %>% st_drop_geometry() %>% filter(Region == "GOM")

stom_gom %>% ggplot(aes(x = month, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)

stom_gom <- stom_gom %>% filter(month == 4  & year != 2016)

stom_gom %>% ggplot(aes(x = month, fill = factor(month))) + geom_histogram(stat = "count") + facet_wrap(~ year)

stom_gom <- stom_gom %>% mutate(date = make_date(year, month, day))
#what is their peak timing of abundance each year?
stom_gom$jd <- yday(stom_gom$date)

stom_ct <- stom_gom %>% mutate(mx = jd*pyamtw)

stom_ct <- stom_ct %>% ungroup %>% group_by(year, analsci) %>% mutate(ct = (sum(mx))/(sum(pyamtw)), avg_abd = mean(pyamtw), sum_abd = sum(pyamtw))

stom_ct <- stom_ct %>% select(year, analsci, ct, avg_abd, sum_abd) %>% unique()


stom_ct$ct_day <- stom_ct$ct - 31 - 29 - 31

stom_ct %>% ggplot(aes(x = year, y = ct_day)) + geom_point() + labs(y = "Spring central tendency of abundance (April 1 - May 31)", main = "Gulf of Maine")+ geom_smooth(aes(x = year, y = ct_day), method = "loess", colour = "blue")


sand_ct <- stom_ct %>% filter(analsci == "AMMODYTIDAE")
```

```{r}
whale_ct <- left_join(whale_ct, sand_ct, by = c("year"))


```

```{r}
ggplot(data = whale_ct ,aes(x = ct_day.x, y = ct_day.y)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = ct_day.x, y = ct_day.y), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


```


ecomon 

```{r}
#now sample grid ID for whale table and whale table 
ecomon$season <- ifelse(ecomon$month == 1 | ecomon$month == 2, "JanFeb", ifelse(ecomon$month == 3 | ecomon$month == 4, "MarApril",ifelse(ecomon$month == 5 | ecomon$month == 6, "MayJun",ifelse(ecomon$month == 7 | ecomon$month == 8, "JulAug",ifelse(ecomon$month == 9 | ecomon$month == 10, "SepOct",ifelse(ecomon$month == 11 | ecomon$month == 12, "NovDec", NA))))))

ecomon$month_proxy <- ifelse(ecomon$season == "JanFeb", 1, ifelse(ecomon$season == "MarApril", 2,ifelse( ecomon$season == "MayJun",3, ifelse(ecomon$season == "JulAug",4, ifelse(ecomon$season == "SepOct", 5,ifelse(ecomon$season == "NovDec", 6, NA))))))


ecomon_spat <- ecomon


coordinates(ecomon_spat) <- ~ lon + lat
ecomon_spat <- st_as_sf(ecomon_spat)

ecomon_spat <- st_set_crs(ecomon_spat, 4326)

ecomon_spat <- st_intersection(ecomon_spat,stratum)
#ecomon_extract <- st_join(ecomon_spat, stratum, join = st_nearest_feature)

ggplot()  + geom_sf(data = world)  + geom_sf(data = ecomon_spat, aes(colour = as.factor(Region))) +
    coord_sf(xlim=c(-90, -50), ylim=c(30,50), expand = TRUE)
```

```{r}
ecomon_gom <- ecomon_spat %>% st_drop_geometry() %>% filter(Region == "GOM")



ecomon_gom %>% ggplot(aes(x = month_proxy, fill = factor(month_proxy))) + geom_histogram(stat = "count") + facet_wrap(~ year)

ecomon_gom <- ecomon_gom %>% filter(year > 2000) %>% filter(year!= 2016 & year !=2018)
ecomon_gom %>% ggplot(aes(x = month_proxy, fill = factor(month_proxy))) + geom_histogram(stat = "count") + facet_wrap(~ year)

ecomon_gom <- ecomon_gom %>% filter(month_proxy == 2 | month_proxy == 3)

ecomon_gom <- ecomon_gom %>% mutate(date = make_date(year, month, day))
#what is their peak timing of abundance each year?
ecomon_gom$jd <- yday(ecomon_gom$date)

ecomon_gom <- ecomon_gom %>% pivot_longer(cols = ctyp_10m2:lopame_100m3, names_to = "spp", values_to = "abd")

ecomon_ct <- ecomon_gom %>% mutate(mx = month_proxy*abd)

ecomon_ct <- ecomon_ct %>% ungroup %>% group_by(year, spp) %>% mutate(ct = (sum(mx))/(sum(abd)))

ecomon_ct <- ecomon_ct %>% select(year, spp, ct) %>% unique()



ecomon_ct %>% ggplot(aes(x = year, y = ct)) + geom_point() + labs(y = "Spring central tendency of abundance (April 1 - May 31)", main = "Gulf of Maine")+ geom_smooth(aes(x = year, y = ct), method = "loess", colour = "blue")

ecomon_ct <- ecomon_ct %>% pivot_wider(names_from = spp, values_from = ct)

```

```{r}
whale_ct <- left_join(whale_ct, ecomon_ct, by = c("year"))


```

```{r}
ggplot(data = whale_ct ,aes(x = ct_day, y = ctyp_10m2)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "black")) + geom_smooth(aes(x = ct_day, y = ctyp_10m2), method = "glm", colour = "blue")+
  stat_poly_line() +
  stat_poly_eq(use_label(c("p", "R2"))) + geom_point()


```




